---
title: sabinaNSDM README
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE, eval=FALSE}
knitr::opts_knit$set(root.dir = "../output/sabinaNSDM/READMEexample/")
```

## Summary of main sabinaNSDM functions


| Overall Step            | Function          | Objective                                       |
|-----------------|-------------------|:-----------------:|
| Data preparation| NSDM.InputData    | Provides the package with the species occurrences and environmental covariates at both global and regional scales|
|                 | NSDM.FormattingData| Background data generation and species occurrences (and absences if available) thinning|
|                 | NSDM.SelectCovariates| Selects uncorrelated and the most relevant environmental covariates|
|Single scale modelling | NSDM.Global| Calibrates, evaluates, and projects ensemble models at the global scale|
|     | NSDM.Regional    | Calibrates, evaluates, and projects ensemble models at the regional scale    |
|Nested modelling  | NSDM.Covariate    | Generate spatially-nested hierarchical species distribution models with the covariate approach. The covariate approach uses the output of the global model as an additional covariate for the regional scale model   |
|     | NSDM.Multiply   | Generate spatially-nested hierarchical species distribution models with the multiply approach. The multiply approach averages the global and regional models    |

## Tutorials

-   [Single species modelling](https://besjournals.onlinelibrary.wiley.com/action/downloadSupplement?doi=10.1111%2F2041-210X.14417&file=mee314417-sup-0001-Supinfo1.pdf)
-   [Multispecies modelling](https://besjournals.onlinelibrary.wiley.com/action/downloadSupplement?doi=10.1111%2F2041-210X.14417&file=mee314417-sup-0002-Supinfo2.pdf)

## Example

This is an example on how to use the <strong>sabinaNSDM</strong> package for conducting spatially-nested hierarchical species distribution modelling.

### Data preparation

First, set your working directory and load the required packages.

```{r eval = TRUE}
#setwd("../output/sabinaNSDM/READMEexample/")

library(terra)
library(covsel)
library(biomod2)
library(ecospat)
library(fs)
library(sgsR)

# Load the sabinaNSDM package
library(sabinaNSDM)
```

Define the species name
```{r eval = TRUE}
SpeciesName <- "Fagus.sylvativa"
```

Load species occurrence and environmental covariates data. Species occurrence *data.frame* must include only two columns: “x” and ”y” coordinates. No row names. The coordinate projection must match that used for the covariates.

```{r eval = TRUE}
# Species occurrences
data(Fagus.sylvatica.xy.global, package = "sabinaNSDM")
spp.data.global <- Fagus.sylvatica.xy.global
data(Fagus.sylvatica.xy.regional, package = "sabinaNSDM")
spp.data.regional <- Fagus.sylvatica.xy.regional
```

The covariates for each spatial scale (i.e., global and regional) should be provided as *SpatRaster*, with each band corresponding to a different covariate. The regional-scale *SpatRaster*  must include all the covariates included in the global-scale file, and it can additionally include covariates only available at this level.

```{r eval = TRUE}
data(expl.var.global, package = "sabinaNSDM")
data(expl.var.regional, package = "sabinaNSDM")
expl.var.global <- terra::unwrap(expl.var.global)
expl.var.regional <- terra::unwrap(expl.var.regional)
```

Additionally, regional-scale *SpatRaster* or a *list of SpatRaster objects* corresponding to the covariates used to project the models at one or several different scenarios (i.e., new scenarios) can be provided.  

```{r eval = TRUE}
# new scenarios
data(new.env, package = "sabinaNSDM")
new.env <- terra::unwrap(new.env)
```

Load the required data for the package with the *NSDM.InputData()* function
```{r eval = TRUE}
nsdm_input <- NSDM.InputData(SpeciesName = SpeciesName,
                             spp.data.global = Fagus.sylvatica.xy.global, 
                             spp.data.regional = Fagus.sylvatica.xy.regional, 
                             expl.var.global = expl.var.global, 
                             expl.var.regional = expl.var.regional,
                             new.env = new.env,
                             new.env.names = "scenario1",
                             Background.Global = NULL, 
                             Background.Regional = NULL,
                             Absences.Global = NULL,
                             Absences.Regional = NULL)
```

Format the data with the *NSDM.FormattingData()* function. This function generates random or stratified background points for model calibration when no specific background or true absence data was loaded in the *NSDM.InputData()* function. Additionally, it applies spatial thinning to species occurrences and absences (if available) to remove duplicates and enforce a minimum distance criterion (by default the resolution of the variables). 
```{r eval = TRUE}
nsdm_finput <- NSDM.FormattingData(nsdm_input,
                                   nPoints = 100, # number of background points
                                   Min.Dist.Global = "resolution",
                                   Min.Dist.Regional = "resolution",
                                   Background.method = "random", # method “random" or "stratified” to generate background points 
                                   save.output = FALSE) #save outputs locally
```

*NSDM.SelectCovariates()* function selects the most relevant and uncorrelated environmental covariates for both global and regional scales.
```{r eval = TRUE}
nsdm_selvars <- NSDM.SelectCovariates(nsdm_finput,
                                      maxncov.Global = 5,   # Max number of covariates to be selected at the global scale
                                      maxncov.Regional = 7, # Max number of covariates to be selected at the regional scale
                                      corcut = 0.7, #  correlation threshold
                                      algorithms = c("glm","gam","rf"),
                                      ClimaticVariablesBands = NULL, # covariate bands to be excluded in the covariate selection at the regional scale
                                      save.output = FALSE)

```

### Single scale modelling
*NSDM.Global()* function generates the global component of the NSDM.

```{r eval = TRUE}
nsdm_global <- NSDM.Global(nsdm_selvars,
                           algorithms = c("GAM","GBM", "RF", "MAXNET","GLM"),# Statistical algorithms used for modelling
                           CV.nb.rep = 10, # number of cross-validation repetitions
                           CV.perc = 0.8, # percentage of the data will be used for training in each cross-validation fold
                           metric.select.thresh = 0.8, #  AUC threshold to include replicates in the final ensemble model
                           CustomModelOptions = NULL, # Allows users to apply custom modelling options. 
                           save.output = FALSE, 
                           rm.biomod.folder = TRUE) # Remove the temporary folders created by `biomod2` 

```

*NSDM.Regional()* function generates the regional component of the NSDM.

```{r eval = FALSE}
# nsdm_regional <- NSDM.Regional(nsdm_selvars,
#                 algorithms = c("GAM","GBM", "RF", "MAXNET","GLM"),
#                 CV.nb.rep = 10,
#                 CV.perc = 0.8,
#                 # metric.select.thresh = 0.8,
#                 CustomModelOptions = NULL, 
#                 save.output = FALSE,
#                 rm.biomod.folder = TRUE)
# 
```

### Nested modelling 

*NSDM.Covariate()* function generates a NSDM with the covariate strategy. The covariate strategy incorporates the output of global models as an additional covariate in the regional model. 
```{r eval = TRUE}
nsdm_covariate <- NSDM.Covariate(nsdm_global,
                                 algorithms = c("GAM","GBM", "RF", "MAXNET","GLM"),
                                 rm.corr=TRUE,
                                 CV.nb.rep = 10,
                                 CV.perc = 0.8,
                                 # metric.select.thresh = 0.8,
                                 CustomModelOptions = NULL,
                                 save.output = FALSE,
                                 rm.biomod.folder = TRUE)

summary(nsdm_covariate)
nsdm_covariate$myEMeval.Ensemble

Covariate.model <- terra::rast(nsdm_covariate$current.projections$Pred) 
terra::plot(Covariate.model)
```

*NSDM.Multiply()* function generates a NSDM with the multiply strategy. The covariate averages the output of the global and the regional models. 

```{r eval = FALSE}
# nsdm_multiply <- NSDM.Multiply(nsdm_global,
#                 nsdm_regional,
#                 method = "Arithmetic", # Method for averate model outputs: "Arithmetic" or "Geometric" mean
#                 rescale = FALSE,
#                 save.output=FALSE)
```


<!-- Multiple studies have shown that NSDMs outperform their non- hierarchical counterparts in various applications (*Journal of Vegetation Science*, <https://doi.org/10.1111/jvs.12726>,Mateo et al. 2019; *Frontiers in Ecology and Evolution*, <https://doi.org/10.3389/fevo.2022.944116>,Chevalier et al. 2022; *Ecography*, <https://doi.org/10.1016/j.ecoinf.2023.102080>; Goicolea et al. in press)--> 
