---
title: "Where will the climate still support raised bogs? Projecting the loss of climatic suitability for Norway's typical raised bogs with a three-class distribution model and warm-flank evidence from Europe"
# TODO(JV): working title only.
format:
  plos-pdf:
    number-sections: false
    journal:
      id: plosone
      # TODO(JV): target journal. Notebook 2025-05-14 mentions Global Change Biology; the
      # PLOS template is kept until that is decided.
    # The .bib is a Better BibLaTeX export (date/journaltitle/\mkbibemph), which the
    # extension's default natbib + plos2015.bst route cannot read. Let citeproc parse
    # it and format with the PLOS CSL the extension already sets.
    cite-method: citeproc
    include-in-header:
      text: |
        % \usepackage{setspace}
        % \doublespacing
execute:
  echo: false
author:
  - name: Julien Vollering
    affiliations:
      - ref: aff1
    corresponding: true
    email: julien.vollering@hvl.no
  # TODO(JV): co-authors and order (Anders Lyngstad, Knut Rydgren, Asbjørn Moen?, ...).
  - name: Anders Lyngstad
    affiliations:
      - ref: aff2
  - name: Knut Rydgren
    affiliations:
      - ref: aff1
affiliations:
  - id: aff1
    name: Department of Environmental Sciences, Western Norway University of Applied Sciences
    city: Sogndal
    country: Norway
  - id: aff2
    name: NTNU University Museum, Norwegian University of Science and Technology
    city: Trondheim
    country: Norway
abstract: |
  TODO(JV): abstract (max 300 words). Suggested skeleton: (1) raised bogs are landforms
  sustained by a narrow climatic window, and Norway's typical raised bogs sit at the cool
  end of that window; (2) we pooled a complete aerial-photo inventory of Norwegian raised
  bogs with European raised-bog records and mapped peatland, so the model could learn where
  raised bogs stop on the warm side; (3) a three-class model (raised bog, other peatland,
  non-peat) projects that 96% of mapped Norwegian raised bogs lose climatic suitability by
  2071-2100 under SSP3-7.0; (4) the same conclusion follows without any model, from the
  observed prevalence of raised bogs in the climates the future brings; (5) cross-validation
  organised along the warming axis shows how far the model can be trusted and where it
  cannot.
author-summary: |
  TODO(JV): author summary (150-200 words, first person; not used by PLOS ONE).
bibliography: Repeat-raisedbogDM.bib
---

<!--
DRAFTING NOTES (delete before submission)

* This draft covers Materials and methods and Results only. Introduction and Discussion are
  placeholders. Search for "TODO" for every point that needs input.
* It describes the pipeline as of the full rerun of 2026-09-16 (output/run_manifest.csv;
  R/RUNALL.R: 24 scripts, pl0 + pl2). Retired designs (hierarchical global/local cascade,
  binary response with the AR50 mask, RFQ, separate pl3 cross-validation, DI as the skill
  axis) are deliberately not mentioned.
* Figures and tables are produced by ms/make_figures.R from pipeline outputs. NUMBERS IN THE
  TEXT ARE NOT TRANSCRIBED: Results (and every Methods number that has a source file) are
  inline R, read in the `setup` chunk below from ms/tables/*.csv and from the pipeline CSVs
  listed in output/run_manifest.csv. The setup chunk refuses to render if the latest run of
  any pipeline script failed, if ms/tables is older than the pipeline outputs (rerun
  ms/make_figures.R), or if a qualitative claim the prose relies on no longer holds
  (guards are marked GUARD).
* Methods numbers still typed by hand because no output file carries them (check after any
  rerun of pl0): 1,137 Norwegian bog cells and the 1.76 million absence candidates; the AR50
  label counts (1.32 million non-peat, 61,850 other peatland, 177,860 mixed, 197,190 without
  value); the land-use screen retention percentages; 524 European presence cells. Design
  settings (0.5 / 0.1 AR50 fractions, 25 absences per bog cell, 1 degree strata, 1,000 trees)
  are in output/pipeline_settings.csv and are also typed. The numbers in the YAML abstract
  placeholder are likewise not coded.
* The gsp no-data sentinel (found 2026-09-12) is fixed and the pipeline was rerun on
  2026-09-16: no training or projection row carries it any more (guarded below). The fix
  changed one thing the old prose relied on: gsp now shifts by about one standard deviation
  under the scenario (it was 0.36 SD while the sentinel inflated its SD), so "precipitation
  barely changes" is gone from Methods and Results. The header of R/pl2_exploreOccupancy.R
  still says every precipitation variable shifts by less than 0.4 SD.
-->

```{r}
#| label: setup
#| include: false
# Every number quoted in the prose is read here. Paths are relative to ms/
# (quarto render ms/ms.qmd).
library(readr)
library(dplyr)
library(tidyr)
library(knitr)

OUT <- "../output"
rd <- function(...) read_csv(file.path(...), show_col_types = FALSE)

## GUARD: the pipeline is complete and the manuscript tables are not older than it ####
man <- rd(OUT, "run_manifest.csv")
last_run <- man |> arrange(end) |> group_by(script) |> slice_tail(n = 1) |> ungroup()
stopifnot(
  "latest run of a pipeline script failed (output/run_manifest.csv)" =
    all(last_run$status == "ok")
)
produced <- file.path(OUT, unlist(strsplit(last_run$files[nzchar(last_run$files)], ";")))
produced <- produced[file.exists(produced)]
stopifnot(
  "ms/tables is older than the pipeline outputs: rerun ms/make_figures.R" =
    all(file.mtime(list.files("tables", full.names = TRUE)) >= max(file.mtime(produced)))
)

## Formatting helpers ####
n0 <- function(x) format(round(x), big.mark = ",", trim = TRUE, scientific = FALSE)
f1 <- function(x, d = 1) formatC(x, format = "f", digits = d)
f2 <- function(x) f1(x, 2)
pc <- function(x, d = 0) paste0(formatC(100 * x, format = "f", digits = d), "%")
sg <- function(x, d = 1) formatC(x, format = "f", digits = d, flag = "+")
fp <- function(x) paste0(formatC(x, format = "f", digits = ifelse(x < 1, 2, 1)), "%") # x in percent
cap1 <- function(x) sub("^(.)", "\\U\\1", x, perl = TRUE)

## Tables and CSVs ####
sn <- tibble::deframe(rd("tables", "study_numbers.csv"))
S <- function(k) unname(sn[[k]])
stopifnot(
  "gsp no-data code present in the training or projection frame" =
    S("gsp_nodata_rows_training") == 0 && S("gsp_nodata_rows_future") == 0
)

frame <- rd("tables", "tbl_frame.csv")
fr <- function(resp, col = "total") frame[[col]][frame$response == resp]
n_all <- sum(frame$total)

cv <- rd("tables", "tbl_cv.csv")
cvv <- function(arm, slice, col) cv[[col]][cv$arm == arm & cv$slice == slice]

ci <- rd(OUT, "pl2", "error_profiles_ci.csv")
bn <- function(b, metric, what = "estimate") ci[[what]][ci$bin == b & ci$metric == metric]
bci <- function(b, metric) {
  sprintf("%s (%s--%s)", f2(bn(b, metric)), f2(bn(b, metric, "lower")), f2(bn(b, metric, "upper")))
}
bins <- ci |> distinct(bin, n, n_bog, n_folds, axis_median, axis_lo, axis_hi)
WARM <- max(bins$bin)
cmp <- rd(OUT, "pl2", "error_profiles_by_bin_compare.csv")
bin_of <- function(x) bins$bin[bins$axis_lo <= x & x <= bins$axis_hi][1]
dip <- rd(OUT, "pl2", "error_profiles_by_bin_di.csv")
rng <- function(x) if (min(x) == max(x)) as.character(min(x)) else paste(min(x), "to", max(x))

occ <- rd("tables", "tbl_occupancy.csv")
oc <- function(b, g, col) occ[[col]][occ$bio10_band == b & occ$gsp_band == g]
WET <- "> 650"
fut_wet <- sum(occ$future_bog_pct[occ$gsp_band == WET]) / 100
fut_warm <- sum(occ$future_bog_pct[occ$bio10_band == ">18"]) / 100
occ12 <- rd(OUT, "pl2", "occupancy_grid_bio12.csv")
warm_cells <- rd(OUT, "pl2", "occupancy_warm_evidence_cells.csv")

trans <- rd("tables", "tbl_transitions.csv")
shr <- rd("tables", "tbl_domain_class_share.csv")
sh <- function(scn, cls, col = "majority_share") shr[[col]][shr$scenario == scn & shr$class == cls]
pr <- rd("tables", "tblS_predictors.csv")
shift <- function(v) pr$projected_shift_sd[pr$predictor == v]
oob <- rd(OUT, "pl2", "oob_metrics_production.csv")
ob <- function(cls, col) oob[[col]][oob$class == cls]
rel <- rd(OUT, "pl2", "reliability_summary.csv")
rl <- function(scn, col) rel[[col]][rel$scenario == scn]
relL <- rd(OUT, "pl2", "reliability_at_lyngstad.csv")
rL <- function(scn, col) relL[[col]][relL$scenario == scn]
cov <- rd(OUT, "pl2", "cv_novelty_coverage.csv")
cvcov <- function(q) cov$cv_heldout[cov$arm == "pairwise" & cov$quantile == q]
part <- rd("tables", "tbl_partitions.csv")

## Predictors that shift most (model-free section) ####
q15 <- rd(OUT, "pl2", "occupancy_quantiles_by_group.csv") |>
  select(feature, group, q05, median, q95) |>
  pivot_wider(
    names_from = group, values_from = c(q05, median, q95),
    names_glue = "{.value}_{gsub(' ', '_', group)}"
  )
top15 <- q15$feature
PRECIP <- c(sprintf("bio%02d", 12:19), "gsp")
stopifnot( # GUARD: the prose says gsp is the only precipitation variable among them
  "gsp is no longer the only precipitation variable among the strongly shifting predictors" =
    identical(intersect(top15, PRECIP), "gsp")
)
n_beyond <- sum(
  q15$median_future_bog < q15$q05_current_bog | q15$median_future_bog > q15$q95_current_bog
)
n_in_np <- sum(
  q15$median_future_bog >= q15$q05_current_nonpeat & q15$median_future_bog <= q15$q95_current_nonpeat
)
qv <- function(f, g) q15[[paste0("median_", g)]][q15$feature == f]

## Partitioning and ruler ####
wts <- rd(OUT, "pl2", "weights_feature_data_partitioning.csv") |>
  filter(method == "Balanced Random Forest")
mat <- wts |> filter(material) |> arrange(desc(median))
part_axis <- mat$feature[1]
part_axis_desc <- sub("^(.)", "\\L\\1", pr$description[pr$predictor == part_axis], perl = TRUE)
bio10_rank <- which(mat$feature == "bio10")
n_material <- nrow(mat)
n_climate <- sum(wts$dynamic)
static_share <- sum(wts$median[!wts$dynamic]) / sum(wts$median)
slope_share <- wts$median[wts$feature == "slope"] / sum(wts$median)
stopifnot( # GUARD: the skill curves are indexed on bio10; the prose reports both axes
  "error-profile axis feature is no longer bio10" =
    rd(OUT, "pipeline_settings.csv") |>
      filter(script == "R/pl2_fitErrorProfiles.R", setting == "axis_feature") |>
      pull(value) == "bio10"
)

## Data blocks ####
eu_alloc <- rd(OUT, "pl0", "eu_absence_allocation.csv")
np_warm <- eu_alloc |> filter(response == "nonpeat", warm)
op_warm <- eu_alloc |> filter(response == "otherpeat", warm)
quota_warm <- max(np_warm$drawn)
quota_cool <- max(eu_alloc$drawn[eu_alloc$response == "nonpeat" & !eu_alloc$warm])
full_warm <- np_warm |> filter(drawn == quota_warm)
epm <- rd(OUT, "pl0", "epm_map_cat_summary.csv")
epm_share <- function(lab) epm$share[epm$label == lab]
msk <- tibble::deframe(rd(OUT, "pl0", "eu_mask_summary.csv"))
dom <- rd(OUT, "pl0", "eu_domain_area_by_country.csv")
aud <- rd(OUT, "pl0", "eu_presence_audit.csv")
aud_stat <- function(set, i) {
  as.numeric(strsplit(aud$bio10_min_q25_med_q75_max[aud$set == set], " / ")[[1]][i])
}
aud_col <- function(set, col) aud[[col]][aud$set == set]
```

# Introduction

TODO(JV): Introduction not drafted. Points the Methods and Results are written to connect
to: raised bogs as ombrotrophic landforms with a climatically bounded distribution
[@joostenMiresPeatlandsEurope2017; @gallego-salaBioclimaticEnvelopeModel2010;
@parviainenClimateEnvelopesMire2007]; the "typical raised bog" region of Europe and its
Norwegian outpost [@tannebergerMiresEuropeRegional2021]; the complete Norwegian inventory
from aerial-photo interpretation [@lyngstadFlybildetolkingAvHogmyr2012;
@lyngstadKartleggingAvTypisk2015]; the management question (restoration prioritisation,
@miljodirektoratetPlanRestaureringAv2020); and the modelling problem that a Norwegian-only
model cannot see the warm edge of the niche because Norway does not contain it.

# Materials and methods

## Study system and question

Typical raised bogs are domed, ombrotrophic peat landforms whose surface is fed only by
precipitation. Their distribution in Europe is confined to a band of climate that is cool
enough to limit evaporation and decomposition but productive enough for *Sphagnum* peat
to accumulate; on the cold and wet side they give way to fens, aapa mires and blanket bogs,
and on the warm and dry side to forest, heath and farmland
[@joostenMiresPeatlandsEurope2017; @tannebergerMiresEuropeRegional2021]. Norway holds the
north-western edge of the typical raised bog region, and Norway's raised bogs therefore sit
at the cool end of the climatic band that supports the landform.

We asked which of Norway's mapped typical raised bogs will still have a climate that
supports a raised bog at the end of this century, and where in Norway such a climate will
exist. Two features of the question shape the analysis. First, the future climate of
Norway's raised bogs lies outside the range of climate that Norway's raised bogs occupy
today, so any model trained on Norwegian data alone is asked to extrapolate. We therefore
recruited data from the rest of the typical raised bog region, where raised bogs and their
warm-side neighbours already occur under the climates Norway is moving into. Second, the
loss of a raised bog under warming does not lead to a single alternative state: the site
may remain peatland of another kind, or it may cease to be peatland at all. We therefore
modelled a three-class response rather than presence and absence, so that the two
ecologically opposite non-bog outcomes are kept apart.

The analysis is organised in four parts: assembling the occurrence data and their labels
(this section and the next three), a model-free description of where the future climate of
today's bogs sits relative to what is observed today, one pooled three-class model and its
projection over Norway, and a cross-validation designed to say how the model behaves when
it is pushed into warmer climates than the bogs it learned from.

## Raised-bog occurrence data

### Norway: a complete aerial-photo inventory

Norwegian raised bogs come from a systematic inventory of typical raised bogs carried out
by stereo interpretation of digital aerial photographs at the NTNU University Museum
between 2012 and 2023 [@lyngstadFlybildetolkingAvHogmyr2012;
@lyngstadKartleggingAvTypisk2015; @lyngstadKartleggingAvTypisk2016;
@lyngstadKartleggingAvTypisk2017; @lyngstadKartleggingAvTypisk2021]. Raised bogs are
domed, with easily visible surface patterns, which makes them well suited to
identification from stereo imagery. Every mire massif in the surveyed areas was assessed,
and massifs classified as concentric, eccentric or plateau raised bog were delineated as
polygons. We used all massifs of these three types (mire-massif type "A" in the survey
database) and dissolved adjoining polygons into `r n0(S("n_lyngstad_polygons"))` contiguous raised-bog polygons with a
total area of `r n0(S("lyngstad_area_km2"))` km^2^ (@fig-studyarea). TODO(JV): confirm that ridge raised bog (kanthøgmyr, A6) is
included in type A and should be, and whether atlantic raised bog is excluded by the
region definition alone.

Because the survey was designed to be complete within the areas it covered, the absence of
a mapped raised bog is informative there and nowhere else. We reconstructed the surveyed
area from the aerial-photo project coverages listed in the survey reports (`r S("n_survey_projects")` orthophoto
projects flown 2008--2015, matched image strip by image strip to the reports) plus the
orthophoto-surveyed area given in @lyngstadKartleggingAvTypisk2016. The survey footprint
covers `r n0(S("footprint_km2"))` km^2^ of southern, eastern and central Norway, and all Norwegian absences
were drawn from inside it (@fig-studyarea (B)). TODO(JV): check the footprint area and the count of
projects against the reports; the Hedmark 2012 survey has no strip table
(notebook 2025-11).

Norway was modelled on a 250 m grid in the European equal-area projection (EPSG:3035).
A cell was a raised-bog presence if its centre fell inside a raised-bog polygon (1,137
cells) and an absence candidate if its centre fell inside the survey footprint but not
inside a polygon (1.76 million cells).

### Europe: raised-bog records from the typical raised bog region

European raised-bog presences were taken from the Natura 2000 habitat inventory
[TODO(JV): citation for the Natura 2000 end-2021 spatial dataset], habitat types 7110
(active raised bogs) and 7120 (degraded raised bogs still capable of regeneration), within
the typical raised bog region of Europe (region IV of @tannebergerMiresEuropeRegional2021,
@fig-studyarea (A)). Natura 2000 sites are polygons that often contain a raised bog as one part of a
larger complex, so a site polygon cannot be used as a presence at 250 m resolution without
a rule for where in the polygon the bog is. We used two rules. Sites smaller than 1 km^2^
were used directly. Larger sites were used only where a vegetation-plot record of raised
bog vegetation (EUNIS habitat Q11 in the European Vegetation Archive plot database,
@chytryEUNISHabitatClassification2020) pinned the bog to a location inside the site; each
such plot became one presence point. Only sites with habitat data quality "good" were
used, and sites with representativity "D" (non-significant presence) were excluded.

This is deliberately narrow. Of the good-quality raised-bog sites in the region, `r n0(aud_col("large, discarded", "polygons"))`
larger sites without a plot record (`r n0(aud_col("large, discarded", "area_km2"))` km^2^ of designated area) were not used as
presences. The loss costs statistical power but not climatic coverage: the median
warmest-quarter temperature of the discarded sites (`r f1(aud_stat("large, discarded", 3))` °C) lies between that of the
small sites retained (`r f1(aud_stat("small, retained", 3))` °C) and the large sites with plots (`r f1(aud_stat("large with Q11, retained", 3))` °C), and the retained
small sites span the widest temperature range of the three sets (`r f1(aud_stat("small, retained", 1))`--`r f1(aud_stat("small, retained", 5))` °C; polygon values on the 5 km working grid). The final
European presence set held `r n0(S("n_eu_presence_geometries"))` geometries (`r n0(msk[["presences (small Natura polygons)"]])` small site polygons and `r n0(msk[["presences (EUNIS Q11 in large Natura polygons)"]])` plot points),
falling in 524 grid cells.

The European domain from which absences were drawn was region IV intersected with the
EU-27 (because Natura 2000 exists only in member states, so raised bogs elsewhere in the
region are unmapped rather than absent), with the country coverage of the European
Peatland Map (next section), and with the extent of our climate data, minus Norway. The
domain covers `r n0(sum(dom$area_km2))` km^2^ in `r nrow(dom)` countries (Finland, Sweden, Latvia, Lithuania,
Estonia, Poland, Austria, France, Germany and Czechia; @fig-studyarea (A)). It contains the whole
range of warmest-quarter temperature that Norway is projected to reach by 2071--2100 (see
Results), which is the property the European data were recruited for.

## A three-class response

### Why three classes

A raised bog is a climatic intermediate. Colder and wetter than the raised-bog band lie
other mire types (fens, aapa mires, blanket bogs); warmer and drier lie forest, heath and
farmland on mineral soil. A binary model of raised bog against everything else lumps these
two opposite flanks into one negative class, and the negatives that define the cold edge
of the niche have nothing in common with those that define the warm edge. Under warming
the relevant transition is away from the cold flank and towards the warm one, so a model
that cannot tell the flanks apart cannot say which way a bog is moving. We therefore
labelled every training cell as one of three classes: *raised bog*, *other peatland*, or
*non-peat*. The probability of raised bog is the quantity of interest; the split of the
remaining probability between the other two classes says what a site is predicted to
become. The same three labels were built on both the Norwegian and the European side, from
different sources but by the same logic (@tbl-frame).

### Norway: labels from the AR50 land-resource map

For Norwegian absence candidates we used the peatland class (class 60, myr) of the AR50
land-resource map [TODO(JV): NIBIO AR50 citation and version], rasterised at 50 m and
aggregated to the fraction of each 250 m cell mapped as peatland. Cells with a peatland
fraction of at least 0.5 were labelled *other peatland*; cells with a fraction below 0.1
were labelled *non-peat*; cells in between (mixed cells) and cells with no AR50 value were
excluded from training rather than assigned to either class. Of the 1.76 million absence
candidates, 1.32 million were non-peat, 61,850 other peatland, 177,860 mixed and 197,190
without a value. The peatland fraction was used only as a label and was not a predictor in
the model.

### Europe: labels from the European Peatland Map 2025

For European absence candidates we used the raster European Peatland Map 2025
[@tannebergerPeatlandMapEurope2017; TODO(JV): add Tegetmeyer et al. 2025 EPM2025 technical description to the library], a compilation of national peatland maps, resampled by majority to the
250 m grid within each country. Its map category 0 (no peatland) became *non-peat* and
category 1 (peat-dominated) became *other peatland*. Category 2 (peat in a soil mosaic)
was excluded, exactly as Norway's mixed cells were: the data producers already flag it as
the class where the extent of peat is uncertain. Inside a mapped country, no value means no
peatland; outside the mapped countries, no value means unknown, and those cells were
excluded rather than treated as non-peat. At 250 m the domain was `r pc(epm_share("non-peat"), 1)` non-peat, `r pc(epm_share("other-peat"), 1)` peat-dominated and `r pc(epm_share("excluded (soil mosaic)"), 1)` soil mosaic.

Before absences were drawn, every cell that could be a raised bog was removed from the
European candidate pool with a deliberately generous exclusion mask: all Natura 2000
raised-bog and blanket-bog sites (7110, 7120, 7130) of any data quality, all raised-bog and
blanket-bog vegetation plots (EUNIS Q11 and Q12) buffered by one cell, and all polygons of
the vector European Peatland Map whose national peatland-type attribute contained "bog",
the whole union dilated by one cell. The mask removed `r pc(msk[["mask share of domain"]], 1)` of the domain (`r n0(msk[["mask area (km2)"]])` km^2^).
Its generosity costs almost nothing in absence supply and ensures that a mapped raised
bog that did not qualify as a presence (for example a large site without a plot record)
became unlabelled rather than a false absence.

### One land-use and water screen for both blocks

Cells dominated by cropland, built-up land or open water carry no information about whether
climate can support peat, and their inclusion as non-peat would charge drainage,
cultivation and the sea to warming. We therefore screened every candidate cell of every
class on both blocks with one rule built from the 10 m ESA WorldCover map for 2021
[TODO(JV): WorldCover citation, Zanaga et al.]: a cell was dropped if more than 50% of
its area was cropland plus built-up land, or more than 50% was water plus unmapped (open
sea). The screen was applied before absences were drawn, so the draw was filled from
eligible cells. It kept 98.2% of Norwegian raised-bog cells (1,116 of 1,137), 89.9% of the
Norwegian absence pool, 95.2% of European raised-bog cells (499 of 524) and 79.5% of the
European absence pool. The two blocks fail the screen for different reasons, Norway mostly
on water and Europe mostly on human land cover, which is why the same rule had to be
applied to both: screening Norway alone would have manufactured a warm-flank asymmetry
instead of removing one.

### A climate-stratified absence draw

The two absence classes are far more numerous than raised bogs on both blocks, and the
non-peat pool in particular is dominated by climates far from any raised bog. Drawing
absences at random would fill the training set with cold uplands and leave the warm edge,
where the European data are needed, thin. We therefore drew a fixed budget of 25 absences
per raised-bog cell per absence class from each block (`r n0(fr("non-peat", "Norway"))` per class in Norway, `r n0(fr("non-peat", "Europe"))`
per class in Europe), spread as evenly as possible across 1 °C strata of mean
warmest-quarter temperature (bio10). Strata with fewer candidates than their share were
taken in full and the remainder redistributed. Strata warmer than the warmest Norwegian
training cell (`r f1(S("no_train_bio10_max"))` °C) were guaranteed a minimum share before the general allocation, so
that the European block always contributed labelled cells beyond the Norwegian envelope.
The final pooled training frame held `r n0(n_all)` cells (@tbl-frame).

## Predictors

Each cell carried 43 predictors (S1 Table). Thirty-two were climate variables from CHELSA
V2.1 [@kargerClimatologiesHighResolution2017; @kargerCHELSAV21Technical2021]: the 19
standard bioclimatic variables and 13 of the CHELSA-BIOCLIM+ variables that have future
projections [@brunGlobalClimaterelatedPredictors2022; @kargerBIOCLIMNovelSet2022] (growing
degree days and number of growing days above 0, 5 and 10 °C, growing-season length,
temperature and precipitation, frost-change frequency, snow-cover days, snow water
equivalent and potential net primary productivity). Current climate is the 1981--2010
climatology; future climate is the 2071--2100 climatology under SSP3-7.0 from the
GFDL-ESM4 model. TODO(JV): only one general circulation model and one scenario are used;
notebook 2026-09-12 raises adding a milder scenario. State the choice of GFDL-ESM4.

Two terrain variables, elevation and slope, were derived by aggregating a digital
elevation model to 250 m and then computing slope in degrees. For Norway the source was the
national 50 m terrain model [TODO(JV): Kartverket DTM50 citation]; for Europe, where that
model does not exist, elevation was fetched from AWS terrain tiles at about 90 m with the
elevatr package [TODO(JV): citation] and processed by the same aggregate-then-slope
recipe. Nine variables summarised deglaciation and long-term climate history from the
CHELSA-TraCE21k palaeoclimate series [TODO(JV): TraCE21k citation] sampled every 500 years
from 16 ka to the present: the time a cell has been continuously ice-free land, and the
mean and sum of annual temperature and precipitation over that ice-free period and over
the whole 16 ka. These express the time a site has had to accumulate peat and were the
same in both scenarios.

Climate and palaeoclimate layers were projected from their native 30 arc-second grid to
a 250 m grid for Norway, which is both the training surface for the Norwegian block and the
projection domain. No European surface was built: the European working grid (5 km) serves
only to define the domain and to stratify the draw on bio10, and predictor values for the
European cells were extracted at 250 m cell centres by bilinear interpolation from the same
native climate and palaeoclimate grids (terrain from the 250 m European terrain raster
above), so that both blocks are bilinear samples of the same native field at a 250 m cell
centre. Growing-degree, growing-season-temperature and snow variables that CHELSA leaves
undefined where the threshold is never crossed were set to zero where all other layers had
data. Growing-season precipitation is stored without a no-data tag, so its no-data code was
masked explicitly before resampling and then treated the same way; the pipeline stops if any
value implies a growing-season precipitation above 100,000 mm. No training or projection cell
carries the code in the results reported here.

## Model-free description of the projection problem

Before fitting any model we asked two questions of the data alone. First, where does the
future climate of today's Norwegian raised bogs lie relative to the climates occupied
today by the three classes? Second, among the training cells that already have such a
climate, how common are raised bogs? The second question is an observational reading of
the same evidence the model will be asked to generalise, and it does not depend on a
learner.

For the first question we compared the 15 predictors that shift most under the scenario
between the three current classes and the `r n0(fr("raised bog", "Norway"))` Norwegian raised-bog cells carried to their
2071--2100 climate. All but one of the 15 are temperature, degree-day, season-length or
temperature-driven productivity measures; the exception is growing-season precipitation
(gsp), which shifts by `r f2(shift("gsp"))` standard deviations of its current spread, while annual precipitation
shifts by only `r f2(shift("bio12"))`. For the
second question we cross-tabulated the training cells on warmest-quarter temperature
(bio10; < 14, 14--16, 16--18 and > 18 °C) and gsp (< 500,
500--650 and > 650 mm) and reported the observed prevalence of raised bog in each cell of
the grid alongside the share of future raised-bog cells that land there. Temperature is
the axis the projection travels. Growing-season precipitation is the moisture axis, chosen
because it is a dynamic predictor tied to the growing season; it is not held constant by the
scenario, so the grid reads prevalence at the moisture the future brings, not at today's.
Raised bogs are ombrotrophic, so a warm-and-dry
analogue would say little about a warm-and-wet future, and the grid keeps the two apart.
The same grid was repeated with annual precipitation (bio12; < 700, 700--1100, > 1100 mm)
as the moisture axis.

## The model

We fitted one random forest to the pooled three-class frame with the randomForestSRC
package [TODO(JV): Ishwaran & Kogalur citation], using the package's balanced-sampling
recipe for imbalanced classes [@chenUsingRandomForest]: each of 1,000 trees was grown on a
class-balanced bootstrap sample (three times the size of the rarest class, drawn with
class-equalising weights), so that the `r pc(fr("raised bog") / n_all)` of cells that are raised bogs count as much as
either abundant class. The predicted class of a cell was the most probable of the three
(argmax), which for a balanced forest needs no separately tuned threshold. We used default
values for the number of variables tried at each split and the terminal node size; random
forests are insensitive to these at this sample size
[@valaviPredictivePerformancePresenceonly2022; @ludwigAssessingImprovingTransferability2023].
The data block (Norway or Europe) was carried through every table as bookkeeping and was
never a predictor: a forest allowed to split on it could learn "Europe" as a shortcut for
"warm" instead of learning the climate.

We note what the class probabilities are and are not. Because the forest was fitted under
a rebalanced prior, its probability of raised bog is not the probability that a cell is a
raised bog at the prevalence found in the landscape. It is a score that ranks cells and,
together with the two other scores, decides the predicted class. Everything below treats
it as such.

## Evaluating transfer into a warmer climate

### Partitions cut along summer temperature

A random cross-validation would test the model on cells with near-identical neighbours in
training and say nothing about the projection, which moves every Norwegian bog into a
climate warmer than any Norwegian training bog [@meyerPredictingUnknownSpace2021;
@linnenbrinkKNNDMCVKfold2024]. We therefore partitioned the training frame in
climate space along the axis the projection travels. The candidate axis was chosen among
predictors that change under the scenario by at least half a standard deviation of their
current spread (`r n_material` of the `r n_climate` climate predictors qualify) as the one with the highest
importance in the forest; in this run that was `r part_axis` (`r part_axis_desc`), with
warmest-quarter temperature (bio10) ranking `r bio10_rank`th among them. The `r n0(fr("raised bog"))`
raised-bog cells were sorted on `r part_axis` and cut into five equal groups of `r n0(part$n_bog[1])`; the other
cells were assigned to the partition whose bog range on that axis they fell in, with the
cuts between partitions placed midway between adjacent bog groups and the two end
partitions open-ended, so that every cell belongs to exactly one partition and cells warmer
than the warmest training bog are kept, in partition 5, rather than dropped (S2 Table;
TODO(JV): decide whether the partition table goes in the main text or supplement).
TODO(JV): the partitions are cut on `r part_axis`, but the skill curves below and the
reliability map are indexed on a bio10 offset (R/pl2_fitErrorProfiles.R hard-codes bio10 as
the axis "the partitions were cut on"). Decide whether the partition axis should be pinned to
bio10 or the offset axis follow the partition axis; the text reports both as they are.

Two cross-validation arms were run on these partitions with the same learner. The
*pairwise* arm trained on one partition and tested on each of the other four (20 train--test
pairs); it is the harder extrapolation test but trains on a median of `r n0(cvv("pairwise", "all", "median_n_train"))` cells against
the production model's `r n0(n_all)`, so its skill confounds extrapolation with data starvation.
The *leave-one-partition-out* arm trained on four partitions and tested on the fifth (5
folds, median `r n0(cvv("lopo", "all", "median_n_train"))` training cells); it is the arm that resembles the production model.
Reading the two together separates the cost of extrapolating from the cost of a small
training set. The leave-one-partition-out arm is quoted as the estimate of production
skill; the pairwise arm is the one the reliability curves are fitted on, for reasons given
in the next section.

Skill was summarised per fold as the 3 x 3 confusion matrix, per-class recall
(sensitivity), per-class one-vs-rest AUC, the macro AUC (mean over classes) and the macro
G-mean (geometric mean of the three recalls, which falls to zero if any class is
abandoned). All metrics were reported for the whole test partition and separately for the
Norwegian and European test cells, so that a good score on one block cannot mask a poor
score on the other. They were also reported separately for test cells warmer than the
warmest training bog and colder than the coldest on the partition axis, where the only meaningful quantity is
the rate at which the model wrongly calls such a cell a raised bog.

### Skill as a function of warming beyond the training bogs

A single cross-validated score describes the model on average, but the projection is read
at a specific place in climate space: about `r f1(S("lyngstad_offset_future_median"), 0)` °C warmer than the bogs the model learned
from. What matters is the model's skill *there*. Following the logic of pixel-wise
accuracy estimation [@meyerPredictingUnknownSpace2021;
@ludwigAssessingImprovingTransferability2023], we used the held-out predictions of the
pairwise arm as a calibration dataset and expressed each held-out cell's
position as its signed offset in warmest-quarter temperature from the median bio10 of the
raised bogs in that fold's training set. A cell at +2 °C was, for the model that scored it,
2 °C warmer than the typical raised bog it had seen. The signed offset was preferred to a
multivariate distance because the model treats warm-side and cold-side novelty
oppositely, and an unsigned distance pools the two.

The pairwise arm was used for this calibration, rather than the leave-one-partition-out
arm quoted for production skill, for two reasons. It reproduces the extrapolation the
projection makes: a model trained on one partition and tested on a distant one scores
held-out raised bogs far from the bogs it learned from, whereas a model trained on four
partitions has a close analogue for nearly every held-out bog and so places few raised-bog
cells at the offsets where the projection is read. Within the 10th--90th percentile band
of the mapped raised bogs' 2071--2100 offset (`r sg(S("lyngstad_offset_future_q10"))` to `r sg(S("lyngstad_offset_future_q90"))` °C), the pairwise arm holds
`r n0(S("band_bog_rows_pairwise"))` scored held-out raised-bog cells, each from one of four models; the leave-one-partition-out arm holds
`r n0(S("band_bog_rows_lopo"))`. And it replicates: every bin of the axis is scored by `r rng(bins$n_folds)`
independently fitted pairwise models, which is what a fold-level interval requires, against
`r rng(cmp$n_folds)` for the leave-one-partition-out arm. The price is that the
curves describe models trained on a median of `r n0(cvv("pairwise", "all", "median_n_train"))` cells, not the production model's
`r n0(n_all)`, so the expected skill read from them is a floor on the production model's rather
than an estimate of it. The leave-one-partition-out point estimates are drawn alongside
for reference.

Held-out cells were binned into eight bins with equal numbers of raised-bog cells (about
800 each; every raised-bog cell is scored once by each of the four models that did not
train on it), so that the warm end of the axis, where the projection sits, is not resting
on a handful of presences, and skill was computed per bin. Because cells within a fold
share one model, uncertainty was estimated by resampling the 20 folds, not cells (cluster
bootstrap, 500 resamples). Every bin draws on `r rng(bins$n_folds)` folds, so an interval is reported
for every bin; the intervals on the operating-point metrics are wide, and should be read
as such.

### Multivariate novelty and area of applicability

Warming along one axis is not the only way a projected cell can differ from the training
data. To capture novelty in all climate dimensions at once we computed the dissimilarity
index of @meyerPredictingUnknownSpace2021: the Euclidean distance from a cell to its
nearest training cell in standardised, importance-weighted predictor space, divided by the
mean distance between training cells. Three choices made this comparable everywhere it was
measured. The weights (median permutation importance over five fits of the production
learner to the whole training frame, 500 trees each) and the standardisation were frozen
once and reused for every fold and for the projection. The distance was computed over the 32 climate
predictors only: terrain and palaeoclimate are identical in both scenarios, cannot register
any projected change, and would only dilute the distance (they carry `r pc(static_share)` of the
unrestricted importance mass, slope alone `r pc(slope_share)`). And the area of applicability threshold
was derived from held-out cross-validation distances (the Tukey upper fence of the
pairwise arm's held-out distances), as in the original method, rather than from
within-training distances, which with `r n0(round(n_all, -3))` densely packed cells measure training density
rather than attainable novelty. The dissimilarity index was used for the applicability
mask and as a second, retained axis for the skill curves; the map's expected skill was
read on the signed temperature offset.

## Projection and reliability mapping

The production forest, fitted on all `r n0(n_all)` cells, was applied to every 250 m land cell of
mainland Norway under current and future climate, giving three class-probability layers
per scenario. Change at the mapped raised bogs was summarised per polygon as the mean of
each layer over the polygon's cells, the predicted class before and after, and the change
in the raised-bog score. For every projected cell we also computed its signed bio10 offset
from the production model's training bogs (median `r f2(S("axis_ref_bio10"))` °C), its dissimilarity index, and
the cross-validated recall and false-positive rate for raised bog interpolated from the
offset bins to the cell's offset; cells beyond the last bin inherit that bin's values,
and cells beyond the applicability threshold are flagged so that no estimate is read
there.

## Software

All analyses were run in R [TODO(JV): version] with the packages sf and terra for
spatial data [TODO(JV): citations], randomForestSRC for the forest, and the tidyverse for
data handling. The full pipeline is available at
<https://github.com/julienvollering/DMraisedbog>. TODO(JV): archive a release (Zenodo)
and cite it.

# Results

## Training data

The pooled training frame held `r n0(fr("raised bog"))` raised-bog cells, `r n0(fr("raised bog", "Norway"))` Norwegian and `r n0(fr("raised bog", "Europe"))` European,
against 25 absences per bog cell in each of the two other classes (@tbl-frame). The Norwegian
raised bogs occupy warmest-quarter temperatures of `r f1(S("no_bog_bio10_min"))`--`r f1(S("no_bog_bio10_max"))` °C (median `r f1(S("no_bog_bio10_median"))` °C), and no
Norwegian training cell of any class is warmer than `r f1(S("no_train_bio10_max"))` °C. The European presences span
`r f1(S("eu_bog_bio10_min"))`--`r f1(S("eu_bog_bio10_max"))` °C and so extend the warm edge of the raised-bog class by `r f1(S("eu_bog_bio10_max") - S("no_bog_bio10_max"))` °C, and the
European absences reach `r f1(S("eu_absence_bio10_max"), 0)` °C. Of the `r n0(sum(np_warm$available))` screened European absence candidates warmer
than any Norwegian training cell, the stratified draw placed `r n0(quota_warm)` non-peat cells in each
1 °C stratum from `r min(full_warm$bio10_from)` to `r max(full_warm$bio10_to)` °C, against `r n0(quota_cool)` per stratum in the cooler strata; other
peatland is scarce that warm (`r n0(sum(op_warm$available))` candidates above `r min(op_warm$bio10_from)` °C, `r n0(sum(op_warm$available[op_warm$bio10_from >= 18]))` above 18 °C), and the draw
took `r n0(sum(op_warm$drawn))` of them.

```{r}
#| label: tbl-frame
#| tbl-cap: "**Composition of the pooled training frame** (250 m cells, current climate). Absences were drawn at 25 per raised-bog cell per class from each block, stratified on warmest-quarter temperature, after the land-use and water screen."
tf <- frame |>
  mutate(response = factor(response, levels = c("raised bog", "other peatland", "non-peat"))) |>
  arrange(response) |>
  mutate(response = as.character(response))
tf <- bind_rows(
  tf,
  tibble(response = "All", Norway = sum(tf$Norway), Europe = sum(tf$Europe), total = sum(tf$total))
)
tf |>
  mutate(across(-response, n0)) |>
  setNames(c("Class", "Norway", "Europe", "Total")) |>
  kable(align = "lrrr")
```

![**Study area.** (A) The European absence domain (green: mire region IV within the
EU-27, the countries mapped by the European Peatland Map, and the climate-data extent) with
the `r n0(S("n_eu_presence_geometries"))` European raised-bog presences (green points), and the Norwegian survey footprint
(orange) with the `r n0(S("n_lyngstad_polygons"))` mapped raised bogs (orange points). (B) Norway with the survey
footprint (black outline) and the mapped raised bogs over mean temperature of the warmest
quarter, 1981--2010.](figures/fig_studyarea.png){#fig-studyarea}

## Where the future takes today's raised bogs, and what is observed there

Under SSP3-7.0 the mean warmest-quarter temperature at the `r n0(fr("raised bog", "Norway"))` Norwegian raised-bog
cells rises from a median of `r f1(S("no_bog_bio10_median"))` °C to `r f1(S("future_bog_bio10_median"))` °C (range `r f1(S("future_bog_bio10_min"))`--`r f1(S("future_bog_bio10_max"))` °C), so that `r pc(S("future_bog_share_above_pooled_bog_max"))` of them
become warmer than the warmest raised bog in the pooled training data (`r f1(S("pooled_bog_bio10_max"))` °C, a European
one). For `r n_beyond` of the 15 most strongly shifting predictors the median future bog cell lies beyond
the central 90% of today's raised bogs, and for all `r n_in_np` of them it lies within the central 90% of today's non-peat cells
(growing degree days above 5 °C, for example, from a median of `r n0(qv("gdd5", "current_bog"))` in today's raised bogs to `r n0(qv("gdd5", "future_bog"))`).
Precipitation does not stay put either: growing-season
precipitation shifts by `r f2(shift("gsp"))` standard deviations (median `r n0(qv("gsp", "current_bog"))` mm to `r n0(qv("gsp", "future_bog"))` mm) and annual precipitation by `r f2(shift("bio12"))`.

The climates the bogs are moving into are already sampled, by the European block. In the
wet column of the occupancy grid (growing-season precipitation above 650 mm), which holds
`r pc(fut_wet)` of the future raised-bog cells, the observed prevalence of raised bog is `r fp(oc("14-16", WET, "prevalence_pct"))` at
14--16 °C, `r fp(oc("16-18", WET, "prevalence_pct"))` at 16--18 °C and `r fp(oc(">18", WET, "prevalence_pct"))` at temperatures above 18 °C (`r oc(">18", WET, "n_bog")` raised bogs among
`r n0(oc(">18", WET, "n_cells"))` cells; @fig-occupancy (B), @tbl-occupancy). `r pc(fut_warm)` of the future raised-bog cells fall in that
warmest band, `r pc(oc(">18", WET, "future_bog_pct") / 100)` of all of them in its wet column: a `r f1(oc("16-18", WET, "prevalence_pct") / oc(">18", WET, "prevalence_pct"), 0)`-fold drop in observed
prevalence from the adjacent band at constant moisture, on `r n0(oc(">18", WET, "n_cells"))` sampled cells, with the moisture that raised bogs
need still present. The grid on annual precipitation shows the same pattern (`r fp(100 * occ12$prevalence[occ12$temp_band == ">18" & occ12$moist_band == "wet >1100"])` raised
bog above 18 °C in the wettest column; supplementary table TODO(JV)).

A caveat belongs with this table. The warm evidence is entirely European and
geographically concentrated: the `r n0(S("warm18_n"))` training cells warmer than 18 °C lie in `r nrow(warm_cells)`
one-degree squares between `r min(warm_cells$grid_lat)` and `r max(warm_cells$grid_lat) + 1` °N (the Massif Central, Jura, Alpine foothills and
the Bohemian belt), with `r pc(warm_cells$cumshare[3])` of them in three squares, at elevations of `r n0(S("warm18_elevation_min"))`--`r n0(S("warm18_elevation_max"))` m. Matching
on temperature and precipitation does not match day length or evaporative demand, and the
effective sample size is far below the cell count.

```{r}
#| label: tbl-occupancy
#| tbl-cap: !expr 'sprintf("**Observed prevalence of raised bog across the climate grid, and where the future raised-bog cells land.** Training cells (current climate) are cross-tabulated on mean temperature of the warmest quarter (bio10) and growing-season precipitation (gsp). The last column is the share of the %s Norwegian raised-bog cells whose 2071--2100 climate falls in that cell of the grid.", format(fr("raised bog", "Norway"), big.mark = ","))'
occ |>
  transmute(
    `bio10 (°C)` = sub("-", "–", bio10_band),
    `gsp (mm)` = sub("-", "–", gsp_band),
    Cells = n0(n_cells),
    `Raised bog` = n0(n_bog),
    Prevalence = sprintf("%.2f%%", prevalence_pct),
    `Share from Europe` = sprintf("%.0f%%", share_eu_pct),
    `Future raised-bog cells` = sprintf("%.1f%%", future_bog_pct)
  ) |>
  kable(align = "llrrrrr")
```

![**The climate the projection moves into, and what is observed there.** (A) Training
cells in the space of warmest-quarter temperature and growing-season precipitation
(15,000 cells drawn at random from each absence class; all `r n0(fr("raised bog"))` raised-bog cells), with the
`r n0(fr("raised bog", "Norway"))` Norwegian raised-bog cells at their 2071--2100 climate (red). Dotted lines are the
bands of panel B. (B) Observed prevalence of raised bog per band, with the number of raised-bog cells
over the number of training cells, and the share of the future raised-bog cells that land
in each band.](figures/fig_occupancy.png){#fig-occupancy}

## Model fit and cross-validated skill

On its own out-of-bag cells the production forest recovered `r pc(ob("bog", "recall"))` of raised bogs (one-vs-rest
AUC `r f2(ob("bog", "auc"))`), `r pc(ob("otherpeat", "recall"))` of other peatland and `r pc(ob("nonpeat", "recall"))` of non-peat, with a macro G-mean of `r f2(oob$Gmean_macro[1])`. As
expected for a balanced forest, the raised-bog class was over-called: `r n0(ob("bog", "n_predicted"))` cells were
predicted raised bog for `r n0(ob("bog", "n_true"))` actual, a precision of `r f2(ob("bog", "precision"))`.

Held-out skill was lower and depended strongly on how much the training set resembled the
production model (@tbl-cv). Leave-one-partition-out folds, with training sets of the
production model's size, reached a macro AUC of `r f2(cvv("lopo", "all", "macro_auc"))` and a macro G-mean of `r f2(cvv("lopo", "all", "macro_gmean"))`, recovering
`r pc(cvv("lopo", "all", "recall_bog"))` of raised bogs; pairwise folds, trained on a median of `r n0(cvv("pairwise", "all", "median_n_train"))` cells (`r pc(cvv("pairwise", "all", "median_n_train") / n_all)` of the production set), reached `r f2(cvv("pairwise", "all", "macro_auc"))` and `r f2(cvv("pairwise", "all", "macro_gmean"))`,
recovering `r pc(cvv("pairwise", "all", "recall_bog"))`. Most of the difference between these arms is the cost of the small training
set rather than of extrapolating, so we quote the leave-one-partition-out arm as the
estimate of production skill under climate transfer.

Skill differed between the two data blocks in a specific way. Raised bogs were recovered
well on both (`r pc(cvv("lopo", "NO", "recall_bog"))` in Norway, `r pc(cvv("lopo", "EU", "recall_bog"))` in Europe), but Norwegian other peatland was
recovered at only `r pc(cvv("lopo", "NO", "recall_otherpeat"))` against `r pc(cvv("lopo", "EU", "recall_otherpeat"))` in Europe, and European non-peat at `r pc(cvv("lopo", "EU", "recall_nonpeat"))` against
`r pc(cvv("lopo", "NO", "recall_nonpeat"))` in Norway (@tbl-cv; S3 Table TODO confusion by block). Slope is by far the most important predictor in the
production forest (permutation importance `r f2(max(pr$importance_all))` against `r f2(sort(pr$importance_all, decreasing = TRUE)[2])` for the next; S1 Table), and
European mapped peatland is flatter than Norway's: `r pc(S("otherpeat_flat_share_EU"))` of
European other-peatland cells have a slope below 1° against `r pc(S("otherpeat_flat_share_NO"))` of Norwegian ones,
so the forest has plausibly learned a "peat is flat"
rule from the European block that it misapplies to Norwegian sloping mires. This weakness
lies on the flat-versus-steep axis, not on the climatic axis the study turns on.

Beyond the temperature range of the training bogs the model was asymmetric, and the
asymmetry is the most informative single result of the cross-validation. When the warmest
partition was held out (`r n0(cvv("lopo", "above", "n_test_rows"))` test cells warmer than every training bog on the partition axis, none of them a
raised bog), the model called `r pc(cvv("lopo", "above", "fpr_bog"))` of them raised bog: a false-positive rate of `r f2(cvv("lopo", "above", "fpr_bog"))`
(pairwise arm `r f2(cvv("pairwise", "above", "fpr_bog"))`). When the coldest partition was held out (`r n0(cvv("lopo", "below", "n_test_rows"))` cells colder than
every training bog) it called `r pc(cvv("lopo", "below", "fpr_bog"))` of them raised bog (pairwise arm `r pc(cvv("pairwise", "below", "fpr_bog"))`). So the forest, asked about a
climate warmer than any raised bog it has seen, does not withdraw the raised-bog label; it
extends it. The projected loss reported below is therefore not an artefact of a model that
defaults to absence outside its training range. It is produced by the warm-flank labels
that are present in the production training set.

```{r}
#| label: tbl-cv
#| tbl-cap: "**Cross-validated skill on partitions cut along the top materially shifting predictor, by arm and by subset of the test cells** (means over folds). Recall is the share of cells of a class predicted as that class; FPR is the share of cells not of the class predicted as it. The two envelope rows contain no raised bogs, so raised-bog recall and the G-mean are undefined there and the false-positive rate is the informative column."
lab_slice <- c(
  all = "all", NO = "Norway", EU = "Europe",
  above = "warmer than all training bogs", below = "colder than all training bogs"
)
f2na <- function(x) ifelse(is.na(x), "--", f2(x))
cv |>
  filter(slice != "inside") |>
  mutate(
    arm = factor(arm, c("lopo", "pairwise")),
    slice = factor(slice, names(lab_slice))
  ) |>
  arrange(arm, slice) |>
  mutate(
    env = slice %in% c("above", "below"),
    Arm = ifelse(duplicated(arm), "", c(lopo = "Leave-one-out", pairwise = "Pairwise")[as.character(arm)]),
    `Test cells` = unname(lab_slice[as.character(slice)]),
    Folds = n_folds,
    `Median training cells` = n0(median_n_train),
    `Macro AUC` = f2na(macro_auc),
    `Macro G-mean` = ifelse(env, "--", f2na(macro_gmean)),
    `Recall raised bog` = ifelse(env, "--", f2na(recall_bog)),
    `Recall other peatland` = f2na(recall_otherpeat),
    `Recall non-peat` = f2na(recall_nonpeat),
    `FPR raised bog` = f2na(fpr_bog)
  ) |>
  select(Arm, `Test cells`, Folds, `Median training cells`, `Macro AUC`, `Macro G-mean`,
         `Recall raised bog`, `Recall other peatland`, `Recall non-peat`, `FPR raised bog`) |>
  kable(align = "llrrrrrrrr")
```

## Skill against warming beyond the training bogs

Along the signed temperature offset, discrimination and the operating point behaved
differently (@fig-skill, @tbl-bins). The raised-bog AUC stayed within `r f2(min(ci$estimate[ci$metric == "auc_bog"]))`--`r f2(max(ci$estimate[ci$metric == "auc_bog"]))` across the eight bins and
did not fall in the warm bin (`r bci(WARM, "auc_bog")`; +`r f1(bins$axis_lo[WARM])` to `r sg(bins$axis_hi[WARM])` °C, median
`r sg(bins$axis_median[WARM])` °C), which draws on `r bins$n_folds[WARM]` folds; the macro AUC did,
from `r f2(max(ci$estimate[ci$metric == "macro_auc" & ci$bin < WARM]))` at its best to `r bci(WARM, "macro_auc")`. The operating point moved
the other way. In the cold bin and within a degree of
the typical training bog (bins 1--5) the pairwise models recovered `r pc(min(ci$estimate[ci$metric == "recall_bog" & ci$bin <= 5]))`--`r pc(max(ci$estimate[ci$metric == "recall_bog" & ci$bin <= 5]))` of raised bogs at
false-positive rates of `r f2(min(ci$estimate[ci$metric == "fpr_bog" & ci$bin <= 5]))`--`r f2(max(ci$estimate[ci$metric == "fpr_bog" & ci$bin <= 5]))`; from `r sg(bins$axis_lo[6])` °C upward recall rose through `r f2(bn(6, "recall_bog"))` and `r f2(bn(7, "recall_bog"))`
to `r bci(WARM, "recall_bog")` in the warm bin, and the false-positive rate through `r f2(bn(6, "fpr_bog"))` and `r f2(bn(7, "fpr_bog"))` to
`r bci(WARM, "fpr_bog")`. The intervals of the warm bin are wide: with `r bins$n_folds[WARM]` folds
they span most of the unit range for both rates. The leave-one-partition-out reference, whose models saw four
partitions, shows the same rise and fall: it recovered `r pc(cmp$recall_bog[1])` of raised bogs in the cold
bin at a false-positive rate of `r f2(cmp$fpr_bog[1])`, `r pc(min(cmp$recall_bog[4:7]))`--`r pc(max(cmp$recall_bog[4:7]))` between `r sg(cmp$axis_lo[4])` and `r sg(max(cmp$axis_median[7], cmp$axis_lo[7]))` °C
(bins 4--7), and `r pc(cmp$recall_bog[WARM])` in its warm bin at a false-positive rate of `r f2(cmp$fpr_bog[WARM])`. On both arms,
then, the false-positive rate is highest in the warm bin: the forest asked about a
climate warmer than its training bogs extends the raised-bog label rather than withholding
it, consistent with the envelope result above. The production-sized reference does not
recover more of the warm bin's raised bogs than the pairwise models do; the two arms differ
mainly in how sharply they separate the interior bins.

The `r n0(S("n_lyngstad_polygons"))` mapped raised bogs sit today at a median offset of `r sg(S("lyngstad_offset_current_median"))` °C, in bin `r bin_of(S("lyngstad_offset_current_median"))`,
where the pairwise models recover `r pc(bn(bin_of(S("lyngstad_offset_current_median")), "recall_bog"))` of the raised bogs and the production-sized reference
`r pc(cmp$recall_bog[bin_of(S("lyngstad_offset_current_median"))])`. Under the 2071--2100 climate they sit at a median offset of `r sg(S("lyngstad_offset_future_median"))` °C (90th
percentile `r sg(S("lyngstad_offset_future_q90"))` °C), inside the warm bin; `r pc(S("lyngstad_share_beyond_bog_edge"))` of the polygons
lie beyond the warmest raised bog in the production training set (`r sg(S("bog_edge_offset"))` °C). The skill the map
can attach to them is the warm bin's: a recall of `r bci(WARM, "recall_bog")` and a false-positive
rate of `r bci(WARM, "fpr_bog")`, from `r bins$n_folds[WARM]` folds.

The same curves against the multivariate dissimilarity index are not monotonic in
discrimination (raised-bog AUC `r f2(dip$auc_bog[1])` in the least novel bin, `r f2(min(dip$auc_bog))` at its lowest and `r f2(dip$auc_bog[nrow(dip)])` in the most
novel, across a `r f1(max(dip$axis_median) / min(dip$axis_median), 0)`-fold range of novelty) while recall falls from `r f2(dip$recall_bog[1])` to `r f2(min(dip$recall_bog))`
and the false-positive rate from `r f2(dip$fpr_bog[1])` to `r f2(dip$fpr_bog[nrow(dip)])`
(S1 Fig). That axis is dominated by cold and seasonally unusual cells rather than
warm ones, which is why it cannot carry the skill read for a warming projection; it is
retained for the applicability mask.

![**Cross-validated skill as a function of warming beyond the training raised
bogs.** Held-out cells of the pairwise arm, binned on their warmest-quarter
temperature offset from the median of the raised bogs in that fold's training set (eight
bins with equal numbers of raised-bog cells; points at bin medians). (A) Recall and
false-positive rate for raised bog at the argmax operating point. (B) One-vs-rest AUC for
raised bog and macro AUC. Ribbons are 95% cluster-bootstrap intervals over the 20 folds.
Thin dashed lines: point estimates of the leave-one-partition-out arm, whose training sets
are the production model's size, for reference. Solid vertical line: the typical training
bog; dashed: the warmest training bog. Rug: the `r n0(S("n_lyngstad_polygons"))` mapped raised bogs at their current
and 2071--2100 offsets; the shaded band spans the 10th--90th percentile of their future
offset.](figures/fig_skill_offset.png){#fig-skill}

```{r}
#| label: tbl-bins
#| tbl-cap: "**Held-out skill by bin of warmest-quarter temperature offset from the training raised bogs** (pairwise arm). Intervals are 95% cluster-bootstrap intervals over the 20 folds."
ci |>
  filter(metric %in% c("recall_bog", "fpr_bog", "auc_bog", "macro_auc")) |>
  mutate(cell = ifelse(
    is.na(lower), f2(estimate),
    sprintf("%s (%s–%s)", f2(estimate), f2(lower), f2(upper))
  )) |>
  select(bin, n, n_bog, n_folds, axis_lo, axis_hi, metric, cell) |>
  pivot_wider(names_from = metric, values_from = cell) |>
  transmute(
    Bin = bin,
    `Offset range (°C)` = sprintf("%+.1f to %+.1f", axis_lo, axis_hi),
    Cells = n0(n),
    `Raised bogs` = n0(n_bog),
    Folds = n_folds,
    `Recall raised bog` = recall_bog,
    `FPR raised bog` = fpr_bog,
    `AUC raised bog` = auc_bog,
    `Macro AUC` = macro_auc
  ) |>
  kable(align = "rlrrrllll")
```

## Projected change at the mapped raised bogs

Under current climate the model reproduces the mapped distribution: the raised-bog score
averages `r f2(S("lyngstad_pbog_current_mean"))` over the `r n0(sum(trans$n_polygons))` polygons and `r n0(sum(trans$n_polygons[trans$class_current == "raised bog"]))` of them are predicted raised bog (@fig-projection (A, C);
@fig-lyngstad (A)). Under the 2071--2100 climate the score averages `r f2(S("lyngstad_pbog_future_mean"))`, it falls at `r pc(S("lyngstad_share_declining"))` of the
polygons (median change `r f2(S("lyngstad_pbog_change_median"))`; 5th and 95th percentiles `r f2(S("lyngstad_pbog_change_q05"))` and `r f2(S("lyngstad_pbog_change_q95"))`), `r ifelse(S("lyngstad_future_pbog_above_0.5") == 0, "no polygon keeps a score above 0.5", pc(S("lyngstad_future_pbog_above_0.5")) |> paste("of polygons keep a score above 0.5"))`, and `r n0(trans$n_polygons[trans$class_current == "raised bog" & trans$class_future == "non-peat"])` of the `r n0(sum(trans$n_polygons[trans$class_current == "raised bog"]))` predicted raised bogs are predicted non-peat, `r n0(trans$n_polygons[trans$class_current == "raised bog" & trans$class_future == "other peatland"])`
other peatland (@tbl-transitions; @fig-projection (B, D); @fig-lyngstad). The direction of the transition is the point
of the three-class response: the mapped raised bogs are not projected to become other
kinds of mire but to leave the peat-supporting climate altogether.

Over mainland Norway as a whole, the share of 2 km cells whose majority class is raised
bog falls from `r pc(sh("current", "bog"), 1)` to `r pc(sh("future", "bog"), 1)`, and other peatland from `r pc(sh("current", "otherpeat"))` to `r pc(sh("future", "otherpeat"))` (TODO(JV): these shares
are computed on 2 km aggregates for plotting; recompute at 250 m if they are to be quoted).
The few cells predicted raised bog in 2071--2100 lie in the far north-east and at scattered
inland sites (@fig-projection (D)). TODO(JV): whether to report "newly suitable" area at all; notebook
2026-06-05 frames the cold-flank transition (other peatland warming into the bog window)
as a second question.

```{r}
#| label: tbl-transitions
#| tbl-cap: !expr 'sprintf("**Predicted class of the %s mapped raised bogs under current and 2071--2100 climate** (argmax of the polygon-mean class probabilities).", format(sum(trans$n_polygons), big.mark = ","))'
trans |>
  transmute(
    `Predicted today` = cap1(class_current),
    `Predicted 2071–2100` = cap1(class_future),
    Polygons = n0(n_polygons),
    Share = sprintf("%.1f%%", share_pct)
  ) |>
  kable(align = "llrr")
```

![**Projected class probabilities over mainland Norway.** Raised-bog score under
(A) 1981--2010 and (B) 2071--2100 climate, and the majority class under (C) 1981--2010 and
(D) 2071--2100. Rasters aggregated to 2 km for display; outlines show the survey
footprint.](figures/fig_projection.png){#fig-projection}

![**Change at the `r n0(S("n_lyngstad_polygons"))` mapped raised bogs.** (A) Polygon-mean raised-bog score under
current against future climate (dotted line: no change). (B) The polygons coloured by their
2071--2100 raised-bog score.](figures/fig_lyngstad_change.png){#fig-lyngstad}

## Reliability of the projection where it is read

By the multivariate dissimilarity index the future projection domain is not extreme: its
median dissimilarity (`r f2(rl("future", "DI_median"))`) is below the median of the held-out cross-validation cells
(`r f2(cvcov("median"))`), its 90th percentile (`r f2(rl("future", "DI_q90"))`) is covered by the cross-validation's 90th percentile
(`r f2(cvcov("q90"))`), and only `r f1(rl("future", "pct_outside_aoa"), 1)`% of the future cells exceed the applicability threshold (`r f2(S("aoa_threshold"))`; maximum `r f2(rl("future", "DI_max"))`), mostly
in the western mountains (@fig-reliability (B)). At the mapped raised
bogs the future dissimilarity is `r f2(rL("future", "DI_median"))`, where the model still has many training neighbours, and `r f1(rL("future", "pct_outside_aoa"), 0)`% of the polygons exceed the threshold.
That reading is optimistic in a specific way: the index measures distance to any training
cell, and the future climate of the bogs is close to European non-peat and other-peatland
cells, not to raised bogs. On the signed temperature axis the same cells sit beyond the
warmest training bog (@fig-reliability (A)). The two layers therefore say different things, and both are
part of the answer: the projection is well supported by labelled data of the two other
classes, and it is an extrapolation with respect to raised bogs specifically. The expected
skill layer (@fig-reliability (C)) carries the warm bin's recall of `r f2(bn(WARM, "recall_bog"))` over the south-eastern
lowlands and the coastal lowlands, and the cold bin's recall of about `r f2(bn(1, "recall_bog"))` over the mountains and the north, so
that the median expected recall over the domain is `r f2(rl("future", "exp_recall_bog_median"))`. At the mapped raised bogs the
expected recall is `r f2(rL("future", "exp_recall_bog"))` (`r f2(rL("future", "exp_recall_bog_lo"))`--`r f2(rL("future", "exp_recall_bog_hi"))`) under the future climate against `r f2(rL("current", "exp_recall_bog"))` (`r f2(rL("current", "exp_recall_bog_lo"))`--`r f2(rL("current", "exp_recall_bog_hi"))`)
today, and the expected false-positive rate `r f2(rL("future", "exp_fpr_bog"))` against `r f2(rL("current", "exp_fpr_bog"))`: the operating point the
projection is read at favours calling raised bog, which makes the non-peat verdict at
these polygons stronger evidence than the same verdict from a neutral operating point
would be. The interval on the future recall is wide, however, and at its lower limit the
forest would miss most raised bogs in that climate, so the verdict is weaker than the point estimate suggests.

![**Reliability layers over the 2071--2100 projection domain.** (A) Warmest-quarter
temperature offset from the production model's training raised bogs. (B) Dissimilarity
index on the frozen ruler (applicability threshold `r f2(S("aoa_threshold"))`; `r f1(rl("future", "pct_outside_aoa"), 1)`% of cells exceed it). (C)
Expected recall of raised bog, read from the offset bins of @fig-skill. Rasters aggregated to
2 km.](figures/fig_reliability.png){#fig-reliability}

# Discussion

TODO(JV): not drafted. Points the Results are written to support: (i) the model-free
occupancy table carries the conclusion and the model adds the map; (ii) the envelope
result (the forest extends the raised-bog label into warmer climates when the warm labels
are withheld) is the reply to "the model just defaults to absence outside its range";
(iii) formation versus persistence: occurrence-based evidence says which climates
maintain raised bogs today, existing bogs are landforms with self-maintaining hydrology,
so the projection is a loss of climatic suitability as currently expressed, not a
disappearance date (notebook 2026-08-21; @joostenMiresPeatlandsEurope2017 p. 135); (iv)
the warm evidence is European, southern and geographically clustered; (v) one scenario and
one climate model; (vi) the other-peatland weakness in Norway is a terrain-source
confound, not a climatic one.

# Supporting information

::: {.supp}
## S1 Table. {#s1-table}

Predictors.

The 43 predictors with their source, whether they change under the scenario (and so enter
the dissimilarity metric), the mean projected shift in standard deviations, and permutation
importance in the production forest (all classes, and for the raised-bog class). Built by
ms/make_figures.R as ms/tables/tblS_predictors.csv.
:::

::: {.supp}
## S2 Table. {#s2-table}

Partitions.

Range on the partition axis (and on warmest-quarter temperature) and class composition of the
five cross-validation partitions (ms/tables/tbl_partitions.csv).
:::

::: {.supp}
## S1 Fig. {#s1-fig}

Cross-validated skill against the dissimilarity index.

Held-out cells of the pairwise arm binned on the dissimilarity index
(eight bins with equal numbers of raised-bog cells; log scale). Ribbons: 95%
cluster-bootstrap intervals over the 20 folds. Dashed line: the applicability threshold. Shaded:
the 10th--90th percentile of the 2071--2100 projection domain (grey) and of the mapped
raised bogs under that climate (red).
:::

# Acknowledgments

TODO(JV).
