---
title: "Where will the climate still support raised bogs? Projecting the loss of climatic suitability for Norway's typical raised bogs with a three-class distribution model and warm-flank evidence from Europe"
# TODO(JV): working title only.
format:
  plos-pdf:
    number-sections: false
    journal:
      id: plosone
      # TODO(JV): target journal. Notebook 2025-05-14 mentions Global Change Biology; the
      # PLOS template is kept until that is decided.
    # The .bib is a Better BibLaTeX export (date/journaltitle/\mkbibemph), which the
    # extension's default natbib + plos2015.bst route cannot read. Let citeproc parse
    # it and format with the PLOS CSL the extension already sets.
    cite-method: citeproc
    include-in-header:
      text: |
        % \usepackage{setspace}
        % \doublespacing
execute:
  echo: false
author:
  - name: Julien Vollering
    affiliations:
      - ref: aff1
    corresponding: true
    email: julien.vollering@hvl.no
  # TODO(JV): co-authors and order (Anders Lyngstad, Knut Rydgren, Asbjørn Moen?, ...).
  - name: Anders Lyngstad
    affiliations:
      - ref: aff2
  - name: Knut Rydgren
    affiliations:
      - ref: aff1
affiliations:
  - id: aff1
    name: Department of Environmental Sciences, Western Norway University of Applied Sciences
    city: Sogndal
    country: Norway
  - id: aff2
    name: NTNU University Museum, Norwegian University of Science and Technology
    city: Trondheim
    country: Norway
abstract: |
  TODO(JV): abstract (max 300 words). Suggested skeleton: (1) raised bogs are landforms
  sustained by a narrow climatic window, and Norway's typical raised bogs sit at the cool
  end of that window; (2) we pooled a complete aerial-photo inventory of Norwegian raised
  bogs with European raised-bog records and mapped peatland, so the model could learn where
  raised bogs stop on the warm side; (3) a three-class model (raised bog, other peatland,
  non-peat) projects that 96% of mapped Norwegian raised bogs lose climatic suitability by
  2071-2100 under SSP3-7.0; (4) the same conclusion follows without any model, from the
  observed prevalence of raised bogs in the climates the future brings; (5) cross-validation
  organised along the warming axis shows how far the model can be trusted and where it
  cannot.
author-summary: |
  TODO(JV): author summary (150-200 words, first person; not used by PLOS ONE).
bibliography: Repeat-raisedbogDM.bib
---


<!--
DRAFTING NOTES (delete before submission)

* This draft covers Materials and methods and Results only. Introduction and Discussion are
  placeholders. Search for "TODO" for every point that needs input.
* It describes the pipeline after the partition axis was pinned to bio10 and the CV and
  reliability scripts were rerun (see notebook.md, 2026-10-05; output/run_manifest.csv).
  Retired designs (hierarchical cascade, binary response with the AR50 mask, RFQ, separate
  pl3 cross-validation, DI as a skill axis, top-importance partition axis) are deliberately
  not mentioned.
* SIMPLIFICATION RULES (2026-10-05). One variant of each analysis in the main text: one
  temperature axis (bio10) for the draw, the partitions and the skill curves; one moisture
  axis (gsp); DI only as an applicability mask; four skill metrics (macro AUC, raised-bog
  recall, raised-bog false-positive rate, and raised-bog AUC on the curve); two CV arms
  (leave-one-partition-out = production skill, pairwise = the replicated offset curve).
  Everything else (per-block skill, envelope slices, bin table, the full occupancy grid,
  the polygon-change scatter) is in the supplement.
* Figures and tables are produced by ms/make_figures.R from pipeline outputs. NUMBERS IN THE
  TEXT ARE NOT TRANSCRIBED: Results (and every Methods number that has a source file) are
  inline R, read in the `setup` chunk from ms/tables/*.csv and from the pipeline CSVs listed
  in output/run_manifest.csv. The setup chunk refuses to render if the latest run of any
  pipeline script failed, if ms/tables is older than the pipeline outputs (rerun
  ms/make_figures.R), or if a qualitative claim the prose relies on no longer holds
  (guards are marked GUARD).
* Methods numbers still typed by hand (check after any rerun of pl0): the land-use screen
  removes about 2% of Norwegian and 5% of European bog cells; design settings (0.5 / 0.1
  AR50 fractions, 25 absences per bog cell, 1 degree strata, 1,000 trees) are in
  output/pipeline_settings.csv. The numbers in the YAML abstract placeholder are not coded.
* Open TODOs that are not about citations are collected in the HTML comments next to the
  text they concern.
-->

```{r}
#| label: setup
#| include: false
# Every number quoted in the prose is read here. Paths are relative to ms/
# (quarto render ms/ms.qmd).
library(readr)
library(dplyr)
library(tidyr)
library(knitr)

OUT <- "../output"
rd <- function(...) read_csv(file.path(...), show_col_types = FALSE)

## GUARD: the pipeline is complete and the manuscript tables are not older than it ####
man <- rd(OUT, "run_manifest.csv")
last_run <- man |>
  arrange(end) |>
  group_by(script) |>
  slice_tail(n = 1) |>
  ungroup()
stopifnot(
  "latest run of a pipeline script failed (output/run_manifest.csv)" = all(
    last_run$status == "ok"
  )
)
produced <- file.path(
  OUT,
  unlist(strsplit(last_run$files[nzchar(last_run$files)], ";"))
)
produced <- produced[file.exists(produced)]
stopifnot(
  "ms/tables is older than the pipeline outputs: rerun ms/make_figures.R" = all(
    file.mtime(list.files("tables", full.names = TRUE)) >=
      max(file.mtime(produced))
  )
)

## Formatting helpers ####
n0 <- function(x) {
  format(round(x), big.mark = ",", trim = TRUE, scientific = FALSE)
}
f1 <- function(x, d = 1) formatC(x, format = "f", digits = d)
f2 <- function(x) f1(x, 2)
pc <- function(x, d = 0) paste0(formatC(100 * x, format = "f", digits = d), "%")
sg <- function(x, d = 1) formatC(x, format = "f", digits = d, flag = "+")
fp <- function(x) {
  paste0(formatC(x, format = "f", digits = ifelse(x < 1, 2, 1)), "%")
} # x in percent
cap1 <- function(x) sub("^(.)", "\\U\\1", x, perl = TRUE)

## Tables and CSVs ####
sn <- tibble::deframe(rd("tables", "study_numbers.csv"))
S <- function(k) unname(sn[[k]])
stopifnot(
  "gsp no-data code present in the training or projection frame" = S(
    "gsp_nodata_rows_training"
  ) ==
    0 &&
    S("gsp_nodata_rows_future") == 0
)
settings <- rd(OUT, "pipeline_settings.csv")
stopifnot(
  # GUARD: partitions, draw and skill curves all sit on bio10; the prose says so
  "partition axis is no longer pinned to bio10" = settings |>
    filter(
      script == "R/pl2_partitionDataByTopFeature.R",
      setting == "axis_rule"
    ) |>
    pull(value) ==
    "pinned: bio10",
  "error-profile axis feature is no longer bio10" = settings |>
    filter(script == "R/pl2_fitErrorProfiles.R", setting == "axis_feature") |>
    pull(value) ==
    "bio10"
)

frame <- rd("tables", "tbl_frame.csv")
fr <- function(resp, col = "total") frame[[col]][frame$response == resp]
n_all <- sum(frame$total)

cv <- rd("tables", "tbl_cv.csv")
cvv <- function(arm, slice, col) cv[[col]][cv$arm == arm & cv$slice == slice]

ci <- rd(OUT, "pl2", "error_profiles_ci.csv")
bn <- function(b, metric, what = "estimate") {
  ci[[what]][ci$bin == b & ci$metric == metric]
}
bci <- function(b, metric) {
  sprintf(
    "%s (%s--%s)",
    f2(bn(b, metric)),
    f2(bn(b, metric, "lower")),
    f2(bn(b, metric, "upper"))
  )
}
bins <- ci |> distinct(bin, n, n_bog, n_folds, axis_median, axis_lo, axis_hi)
WARM <- max(bins$bin)
cmp <- rd(OUT, "pl2", "error_profiles_by_bin_compare.csv")
bin_of <- function(x) bins$bin[bins$axis_lo <= x & x <= bins$axis_hi][1]
rng <- function(x) {
  if (min(x) == max(x)) as.character(min(x)) else paste(min(x), "to", max(x))
}

occ <- rd("tables", "tbl_occupancy.csv")
oc <- function(b, g, col) occ[[col]][occ$bio10_band == b & occ$gsp_band == g]
WET <- "> 650"
fut_wet <- sum(occ$future_bog_pct[occ$gsp_band == WET]) / 100
fut_warm <- sum(occ$future_bog_pct[occ$bio10_band == ">18"]) / 100
warm_cells <- rd(OUT, "pl2", "occupancy_warm_evidence_cells.csv")

trans <- rd("tables", "tbl_transitions.csv")
shr <- rd("tables", "tbl_domain_class_share.csv")
sh <- function(scn, cls, col = "majority_share") {
  shr[[col]][shr$scenario == scn & shr$class == cls]
}
tr_n <- function(cur = NULL, fut = NULL) {
  ok <- rep(TRUE, nrow(trans))
  if (!is.null(cur)) ok <- ok & trans$class_current == cur
  if (!is.null(fut)) ok <- ok & trans$class_future == fut
  sum(trans$n_polygons[ok])
}
pr <- rd("tables", "tblS_predictors.csv")
shift <- function(v) pr$projected_shift_sd[pr$predictor == v]
rel <- rd(OUT, "pl2", "reliability_summary.csv")
rl <- function(scn, col) rel[[col]][rel$scenario == scn]
relL <- rd(OUT, "pl2", "reliability_at_lyngstad.csv")
rL <- function(scn, col) relL[[col]][relL$scenario == scn]
part <- rd("tables", "tbl_partitions.csv")
wts <- rd(OUT, "pl2", "weights_feature_data_partitioning.csv") |>
  filter(method == "Balanced Random Forest")
static_share <- sum(wts$median[!wts$dynamic]) / sum(wts$median)

## Data blocks ####
eu_alloc <- rd(OUT, "pl0", "eu_absence_allocation.csv")
np_warm <- eu_alloc |> filter(response == "nonpeat", warm)
epm <- rd(OUT, "pl0", "epm_map_cat_summary.csv")
epm_share <- function(lab) epm$share[epm$label == lab]
msk <- tibble::deframe(rd(OUT, "pl0", "eu_mask_summary.csv"))
dom <- rd(OUT, "pl0", "eu_domain_area_by_country.csv")
aud <- rd(OUT, "pl0", "eu_presence_audit.csv")
aud_col <- function(set, col) aud[[col]][aud$set == set]
```

# Introduction

TODO(JV): Introduction not drafted. Points the Methods and Results are written to connect
to: raised bogs as ombrotrophic landforms with a climatically bounded distribution
[@joostenMiresPeatlandsEurope2017; @gallego-salaBioclimaticEnvelopeModel2010;
@parviainenClimateEnvelopesMire2007]; the "typical raised bog" region of Europe and its
Norwegian outpost [@tannebergerMiresEuropeRegional2021]; the complete Norwegian inventory
from aerial-photo interpretation [@lyngstadFlybildetolkingAvHogmyr2012;
@lyngstadKartleggingAvTypisk2015]; the management question (restoration prioritisation,
@miljodirektoratetPlanRestaureringAv2020); and the modelling problem that a Norwegian-only
model cannot see the warm edge of the niche because Norway does not contain it.

# Materials and methods

## Study system, question and design

Typical raised bogs are domed, ombrotrophic peat landforms whose surface is fed only by
precipitation. Their distribution in Europe is confined to a band of climate that is cool
enough to limit decomposition but productive enough for *Sphagnum* peat to accumulate; on
the cold and wet side they give way to fens, aapa mires and blanket bogs, and on the warm
and dry side to forest, heath and farmland
[@joostenMiresPeatlandsEurope2017; @tannebergerMiresEuropeRegional2021]. Norway holds the
north-western edge of the typical raised bog region, so Norway's raised bogs sit at the
cool end of the band.

We asked which of Norway's mapped typical raised bogs will still have a climate that
supports a raised bog at the end of this century. The future climate of Norway's raised
bogs lies outside the range they occupy today, so a model trained on Norwegian data alone
would have to extrapolate. The analysis rests on five design decisions, each described in
its own section below.

1. **Pool European and Norwegian data**, so that the model sees raised bogs and their
   warm-side neighbours in the climates Norway is moving into.
2. **Use three classes** (raised bog, other peatland, non-peat), because a bog lost to
   warming becomes either another kind of peatland or no peatland, and the two outcomes
   mean different things.
3. **Build both data blocks the same way**: the same labelling logic, the same land-use
   screen and the same climate-stratified absence draw.
4. **Test transfer along the axis the projection travels**: cross-validation partitions
   and skill curves are both cut on the warmest-quarter temperature (bio10), the same
   axis the absence draw is stratified on.
5. **Read reliability on the signed temperature offset** from the training bogs. A
   multivariate dissimilarity index is used only as an applicability mask.

We first describe the data (this section and the next three), then a model-free
description of the projection problem, one pooled model and its projection, and the
cross-validation that says how the model behaves in warmer climates than its training
bogs.

## Raised-bog occurrence data

### Norway: a complete aerial-photo inventory

Norwegian raised bogs come from a systematic inventory of typical raised bogs carried out
by stereo interpretation of digital aerial photographs at the NTNU University Museum
between 2012 and 2023 [@lyngstadFlybildetolkingAvHogmyr2012;
@lyngstadKartleggingAvTypisk2015; @lyngstadKartleggingAvTypisk2016;
@lyngstadKartleggingAvTypisk2017; @lyngstadKartleggingAvTypisk2021]. Every mire massif in
the surveyed areas was assessed, and massifs classified as concentric, eccentric or plateau
raised bog (mire-massif type "A" in the survey database) were delineated as polygons. We
dissolved adjoining polygons into `r n0(S("n_lyngstad_polygons"))` contiguous raised-bog polygons with a total area of
`r n0(S("lyngstad_area_km2"))` km^2^ (@fig-studyarea).

<!-- TODO(JV): confirm that ridge raised bog (kanthøgmyr, A6) is included in type A and
should be, and whether atlantic raised bog is excluded by the region definition alone. -->

Because the survey was designed to be complete within the areas it covered, the absence of
a mapped raised bog is informative there and nowhere else. We reconstructed the surveyed
area from the aerial-photo project coverages listed in the survey reports (`r S("n_survey_projects")` orthophoto
projects flown 2008--2015) plus the orthophoto-surveyed area given in
@lyngstadKartleggingAvTypisk2016. The survey footprint covers `r n0(S("footprint_km2"))` km^2^ of southern, eastern
and central Norway, and all Norwegian absences were drawn from inside it
(@fig-studyarea (B)).

<!-- TODO(JV): check the footprint area and the project count against the reports; the
Hedmark 2012 survey has no strip table (notebook 2025-11). -->

Norway was modelled on a 250 m grid in the European equal-area projection (EPSG:3035). A
cell was a raised-bog presence if its centre fell inside a raised-bog polygon, and an
absence candidate if its centre fell inside the survey footprint but not inside a polygon.

### Europe: raised-bog records from the typical raised bog region

European raised-bog presences were taken from the Natura 2000 habitat inventory
[TODO cite: Natura 2000 end-2021 spatial dataset], habitat types 7110 (active raised bogs)
and 7120 (degraded raised bogs still capable of regeneration), within the typical raised
bog region of Europe (region IV of @tannebergerMiresEuropeRegional2021, @fig-studyarea
(A)). Natura 2000 sites often contain a raised bog as one part of a larger complex, so a
site polygon cannot be used as a presence at 250 m resolution without a rule for where in
the polygon the bog is. We used two rules. Sites smaller than 1 km^2^ were used directly.
Larger sites were used only where a vegetation-plot record of raised bog vegetation (EUNIS
habitat Q11 in the European Vegetation Archive, @chytryEUNISHabitatClassification2020)
pinned the bog to a location inside the site; each such plot became one presence point.
Only sites with habitat data quality "good" were used, and sites with representativity "D"
(non-significant presence) were excluded. This is deliberately narrow: it discards
`r n0(aud_col("large, discarded", "polygons"))` large sites without a plot record, which costs statistical power but not
climatic coverage (S1 Text). The final European presence set held `r n0(S("n_eu_presence_geometries"))` geometries
(`r n0(msk[["presences (small Natura polygons)"]])` small site polygons and `r n0(msk[["presences (EUNIS Q11 in large Natura polygons)"]])` plot points).

The European domain from which absences were drawn was region IV intersected with the
EU-27 (Natura 2000 exists only in member states, so raised bogs elsewhere in the region are
unmapped rather than absent), with the country coverage of the European Peatland Map (next
section) and with the extent of our climate data, minus Norway. It covers `r n0(sum(dom$area_km2))` km^2^ in `r nrow(dom)`
countries (@fig-studyarea (A)) and contains the whole range of warmest-quarter temperature
that Norway is projected to reach by 2071--2100 (Results), which is the property the
European data were recruited for.

## A three-class response

### Why three classes

A raised bog is a climatic intermediate. Colder and wetter than the raised-bog band lie
other mire types; warmer and drier lie forest, heath and farmland on mineral soil. A binary
model of raised bog against everything else lumps these two opposite flanks into one
negative class, and cannot say which way a bog is moving. We therefore labelled every
training cell as one of three classes: *raised bog*, *other peatland* or *non-peat*. The
probability of raised bog is the quantity of interest; the split of the remaining
probability between the other two classes says what a site is predicted to become. The same
three labels were built on both blocks, from different sources but by the same logic
(@tbl-frame).

### Labels, land-use screen and absence draw

*Norway.* Absence candidates were labelled from the peatland class (class 60, myr) of the
AR50 land-resource map [TODO cite: NIBIO AR50, version], rasterised at 50 m and aggregated
to the fraction of each 250 m cell mapped as peatland. Cells with a peatland fraction of at
least 0.5 were *other peatland*; cells with a fraction below 0.1 were *non-peat*; cells in
between and cells without an AR50 value were excluded rather than assigned to either class.

*Europe.* Absence candidates were labelled from the European Peatland Map 2025
[@tannebergerPeatlandMapEurope2017; TODO cite: EPM2025 technical description], a
compilation of national peatland maps, resampled by majority to the 250 m grid. Map
category 0 (no peatland) became *non-peat* and category 1 (peat-dominated) became *other
peatland*; category 2 (peat in a soil mosaic) was excluded, exactly as Norway's mixed
cells were. Cells outside the mapped countries were excluded as unknown. The domain at
250 m was `r pc(epm_share("non-peat"), 0)` non-peat, `r pc(epm_share("other-peat"), 0)` peat-dominated and `r pc(epm_share("excluded (soil mosaic)"), 0)` soil mosaic. Before absences were drawn,
every cell that could be a raised bog was removed from the European candidate pool with a
deliberately generous mask (all Natura 2000 raised-bog and blanket-bog sites of any quality,
raised-bog and blanket-bog vegetation plots, and mapped bog peatland polygons, dilated by one
cell; `r pc(msk[["mask share of domain"]], 0)` of the domain), so that a mapped raised bog that did not qualify as a
presence became unlabelled rather than a false absence.

*One land-use and water screen.* Cells dominated by cropland, built-up land or open water
carry no information about whether climate can support peat, and their inclusion as
non-peat would charge drainage, cultivation and the sea to warming. Every candidate cell
of every class on both blocks was therefore screened with one rule built from the 10 m ESA
WorldCover map for 2021 [TODO cite: Zanaga et al.]: a cell was dropped if more than 50% of
its area was cropland plus built-up land, or more than 50% was water plus unmapped. The
screen removed about 2% of Norwegian and 5% of European raised-bog cells. It removed more
absence candidates, mostly water in Norway and human land cover in Europe, which is why it
had to be applied to both blocks: screening Norway alone would have manufactured a
warm-flank asymmetry instead of removing one.

*A climate-stratified absence draw.* The two absence classes are far more numerous than
raised bogs, and the non-peat pool in particular is dominated by climates far from any
raised bog. Drawing absences at random would fill the training set with cold uplands and
leave the warm edge, where the European data are needed, thin. We therefore drew 25
absences per raised-bog cell per absence class from each block, spread as evenly as
possible across 1 °C strata of bio10 (mean temperature of the warmest quarter). Strata
with fewer candidates than their share were taken in full and the remainder redistributed.
Strata warmer than the warmest Norwegian training cell (`r f1(S("no_train_bio10_max"))` °C) were guaranteed a minimum
share first, so that the European block always contributed labelled cells beyond the
Norwegian envelope. The pooled training frame held `r n0(n_all)` cells (@tbl-frame).

## Predictors

Each cell carried 43 predictors (S1 Table). Thirty-two were climate variables from CHELSA
V2.1 [@kargerClimatologiesHighResolution2017; @kargerCHELSAV21Technical2021]: the 19
standard bioclimatic variables and 13 CHELSA-BIOCLIM+ variables with future projections
[@brunGlobalClimaterelatedPredictors2022; @kargerBIOCLIMNovelSet2022] (growing degree
days and growing days, growing-season length, temperature and precipitation, frost
frequency, snow cover and productivity). Current climate is the 1981--2010 climatology;
future climate is the 2071--2100 climatology under SSP3-7.0 from the GFDL-ESM4 model. Two
terrain variables (elevation and slope at 250 m) came from the Norwegian 50 m terrain model
[TODO cite: Kartverket DTM50] for Norway and from AWS terrain tiles via the elevatr
package for Europe [TODO cite]. Nine variables summarising time since deglaciation and
long-term climate history came from the CHELSA-TraCE21k palaeoclimate series [TODO cite]
and were the same in both scenarios. Both blocks are bilinear samples of the same native
climate grids at a 250 m cell centre (S1 Text).

<!-- TODO(JV): only one general circulation model and one scenario are used; notebook
2026-09-12 raises adding a milder scenario. State the choice of GFDL-ESM4. -->

## Model-free description of the projection problem

Before fitting any model we asked two questions of the data alone. First, where does the
future climate of today's Norwegian raised bogs lie relative to the climates the training
data sample? Second, among the training cells that already have such a climate, how common
are raised bogs? The second question is an observational reading of the same evidence the
model is asked to generalise, and it does not depend on a learner.

For the first question we carried the `r n0(fr("raised bog", "Norway"))` Norwegian raised-bog cells to their 2071--2100
climate and compared their bio10 with the training data. For the second we cross-tabulated
the training cells on bio10 (< 14, 14--16, 16--18 and > 18 °C) and growing-season
precipitation (gsp; < 500, 500--650 and > 650 mm) and reported the observed prevalence of
raised bog in each cell of the grid next to the share of future raised-bog cells that land
there. Temperature is the axis the projection travels. Growing-season precipitation is the
moisture axis: raised bogs are ombrotrophic, so a warm-and-dry analogue would say little
about a warm-and-wet future, and the grid keeps the two apart. It shifts by `r f2(shift("gsp"))` standard
deviations of its current spread under the scenario, so the grid reads prevalence at the
moisture the future brings, not at today's.

## The model

We fitted one random forest to the pooled three-class frame with the randomForestSRC
package [TODO cite: Ishwaran & Kogalur], using its balanced-sampling recipe for imbalanced
classes [@chenUsingRandomForest]: each of 1,000 trees was grown on a class-balanced
bootstrap sample, so that the `r pc(fr("raised bog") / n_all)` of cells that are raised bogs count as much as either
abundant class. The predicted class of a cell was the most probable of the three (argmax),
which for a balanced forest needs no separately tuned threshold. Default values were used
for the number of variables tried at each split and the terminal node size
[@valaviPredictivePerformancePresenceonly2022; @ludwigAssessingImprovingTransferability2023].
The data block (Norway or Europe) was never a predictor: a forest allowed to split on it
could learn "Europe" as a shortcut for "warm" instead of learning the climate.

Because the forest was fitted under a rebalanced prior, its probability of raised bog is
not the probability that a cell is a raised bog at the prevalence found in the landscape.
It is a score that ranks cells and, together with the two other scores, decides the
predicted class, and we treat it as such.

## Evaluating transfer into a warmer climate

### Partitions and cross-validation arms

A random cross-validation would test the model on cells with near-identical neighbours in
training and say nothing about the projection, which moves every Norwegian bog into a
climate warmer than any Norwegian training bog [@meyerPredictingUnknownSpace2021;
@linnenbrinkKNNDMCVKfold2024]. We therefore partitioned the training frame along bio10.
The `r n0(fr("raised bog"))` raised-bog cells were sorted on bio10 and cut into five equal groups of
`r n0(part$n_bog[1])`; the other cells were assigned to the partition whose bog range they fell in, with the
cuts midway between adjacent bog groups and the two end partitions open-ended, so that
every cell belongs to exactly one partition and cells warmer than the warmest training
bog are kept, in partition 5 (S2 Table).

Two cross-validation arms were run on these partitions with the same learner. The
*leave-one-partition-out* arm trained on four partitions and tested on the fifth (5
folds); it resembles the production model and its skill is quoted as the estimate of
production skill. The *pairwise* arm trained on one partition and tested on each of the
other four (20 train--test pairs). It trains on far fewer cells, so its skill is lower than
the production model's, but it scores raised bogs far from the bogs the model learned from,
and every bin of the offset axis below is scored by several independently fitted models,
which a fold-level interval requires. The pairwise arm is therefore the one the skill
curves are fitted on.

Skill was summarised per fold by the macro AUC (mean over the three one-vs-rest AUCs),
and for raised bog by its recall, its false-positive rate (the share of non-bog cells
predicted raised bog) and its one-vs-rest AUC. Per-class recall, the G-mean and the split
by data block are in S3 Table.

### Skill as a function of warming beyond the training bogs

A single cross-validated score describes the model on average, but the projection is read
at a specific place in climate space: about `r f1(S("lyngstad_offset_future_median"), 0)` °C warmer than the bogs the model learned
from. Following the logic of pixel-wise accuracy estimation
[@meyerPredictingUnknownSpace2021; @ludwigAssessingImprovingTransferability2023], we used
the held-out predictions of the pairwise arm as a calibration dataset and expressed each
held-out cell's position as its signed offset in bio10 from the median bio10 of the raised
bogs in that fold's training set. A cell at +2 °C was, for the model that scored it, 2 °C
warmer than the typical raised bog it had seen. A signed offset was used because the model
treats warm-side and cold-side novelty oppositely, and an unsigned distance pools the two.

Held-out cells were binned into eight bins with equal numbers of raised-bog cells (about
800 each), so that the warm end of the axis, where the projection sits, does not rest on a
handful of presences, and skill was computed per bin. Because cells within a fold share one
model, uncertainty was estimated by resampling the 20 folds, not cells (cluster bootstrap,
500 resamples). The curves describe models trained on a fraction of the production
model's data, so the skill read from them is a floor on the production model's rather than
an estimate of it.

### Area of applicability

Warming along one axis is not the only way a projected cell can differ from the training
data. As an applicability mask we computed the dissimilarity index of
@meyerPredictingUnknownSpace2021: the Euclidean distance from a cell to its nearest
training cell in standardised, importance-weighted predictor space, divided by the mean
distance between training cells. Weights and standardisation were frozen once from the
production fit and reused for every fold and for the projection. Only the 32 climate
predictors entered the distance: terrain and palaeoclimate are identical in both
scenarios, cannot register projected change, and would only dilute the distance (they
carry `r pc(static_share)` of the importance mass). The applicability threshold was the Tukey upper
fence of the pairwise arm's held-out distances, as in the original method. The index was
used only as a mask; expected skill was read on the signed temperature offset.

## Projection and reliability mapping

The production forest, fitted on all `r n0(n_all)` cells, was applied to every 250 m land cell of
mainland Norway under current and future climate, giving three class-probability layers
per scenario. Change at the mapped raised bogs was summarised per polygon as the mean of
each layer over the polygon's cells, the predicted class before and after, and the change
in the raised-bog score. For every projected cell we also computed its signed bio10 offset
from the production model's training bogs (median `r f2(S("axis_ref_bio10"))` °C), its dissimilarity index, and
the cross-validated recall and false-positive rate for raised bog interpolated from the
offset bins to the cell's offset; cells beyond the last bin inherit that bin's values, and
cells beyond the applicability threshold are flagged.

## Software

All analyses were run in R [TODO cite: R version] with the packages sf and terra for
spatial data [TODO cite], randomForestSRC for the forest, and the tidyverse for data
handling. The full pipeline is available at <https://github.com/julienvollering/DMraisedbog>.
TODO(JV): archive a release (Zenodo) and cite it.

# Results

## Training data

The pooled training frame held `r n0(fr("raised bog"))` raised-bog cells, `r n0(fr("raised bog", "Norway"))` Norwegian and `r n0(fr("raised bog", "Europe"))` European,
against 25 absences per bog cell in each of the two other classes (@tbl-frame). The Norwegian
raised bogs occupy warmest-quarter temperatures (bio10) of `r f1(S("no_bog_bio10_min"))`--`r f1(S("no_bog_bio10_max"))` °C (median `r f1(S("no_bog_bio10_median"))` °C), and no
Norwegian training cell of any class is warmer than `r f1(S("no_train_bio10_max"))` °C. The European presences span
`r f1(S("eu_bog_bio10_min"))`--`r f1(S("eu_bog_bio10_max"))` °C and so extend the warm edge of the raised-bog class by `r f1(S("eu_bog_bio10_max") - S("no_bog_bio10_max"))` °C, and the
European absences reach `r f1(S("eu_absence_bio10_max"), 0)` °C, with `r n0(max(np_warm$drawn))` non-peat cells per 1 °C stratum beyond the
Norwegian envelope.

```{r}
#| label: tbl-frame
#| tbl-cap: "**Composition of the pooled training frame** (250 m cells, current climate). Absences were drawn at 25 per raised-bog cell per class from each block, stratified on warmest-quarter temperature, after the land-use and water screen."
tf <- frame |>
  mutate(
    response = factor(
      response,
      levels = c("raised bog", "other peatland", "non-peat")
    )
  ) |>
  arrange(response) |>
  mutate(response = as.character(response))
tf <- bind_rows(
  tf,
  tibble(
    response = "All",
    Norway = sum(tf$Norway),
    Europe = sum(tf$Europe),
    total = sum(tf$total)
  )
)
tf |>
  mutate(across(-response, n0)) |>
  setNames(c("Class", "Norway", "Europe", "Total")) |>
  kable(align = "lrrr")
```

![**Study area.** (A) The European absence domain (green: mire region IV within the
EU-27, the countries mapped by the European Peatland Map, and the climate-data extent) with
the `r n0(S("n_eu_presence_geometries"))` European raised-bog presences (green points), and the Norwegian survey footprint
(orange) with the `r n0(S("n_lyngstad_polygons"))` mapped raised bogs (orange points). (B) Norway with the survey
footprint (black outline) and the mapped raised bogs over mean temperature of the warmest
quarter, 1981--2010.](figures/fig_studyarea.png){#fig-studyarea}

## Where the future takes today's raised bogs, and what is observed there

Under SSP3-7.0 the median bio10 at the `r n0(fr("raised bog", "Norway"))` Norwegian raised-bog cells rises from
`r f1(S("no_bog_bio10_median"))` °C to `r f1(S("future_bog_bio10_median"))` °C (range `r f1(S("future_bog_bio10_min"))`--`r f1(S("future_bog_bio10_max"))` °C), so that `r pc(S("future_bog_share_above_pooled_bog_max"))` of them become
warmer than the warmest raised bog in the pooled training data (`r f1(S("pooled_bog_bio10_max"))` °C, a European one).
Growing-season precipitation shifts too, by `r f2(shift("gsp"))` standard deviations of its current spread, so
moisture is not held constant.

The climates the bogs are moving into are already sampled, by the European block. In the
wet column of the occupancy grid (growing-season precipitation above 650 mm), which holds
`r pc(fut_wet)` of the future raised-bog cells, the observed prevalence of raised bog is `r fp(oc("14-16", WET, "prevalence_pct"))` at
14--16 °C, `r fp(oc("16-18", WET, "prevalence_pct"))` at 16--18 °C and `r fp(oc(">18", WET, "prevalence_pct"))` above 18 °C (@fig-occupancy (B), @tbl-occupancy; the
full grid is S4 Table). `r pc(oc(">18", WET, "future_bog_pct") / 100)` of all future raised-bog cells fall in that last cell: a
`r f1(oc("16-18", WET, "prevalence_pct") / oc(">18", WET, "prevalence_pct"), 0)`-fold drop in observed prevalence from the adjacent band at constant moisture, on
`r n0(oc(">18", WET, "n_cells"))` sampled cells, with the moisture that raised bogs need still present.

The caveat is that the warm evidence is entirely European and geographically concentrated:
the `r n0(S("warm18_n"))` training cells warmer than 18 °C lie in `r nrow(warm_cells)` one-degree squares (the Massif
Central, Jura, Alpine foothills and the Bohemian belt), with `r pc(warm_cells$cumshare[3])` of them in three squares, so
the effective sample size is far below the cell count.

```{r}
#| label: tbl-occupancy
#| tbl-cap: !expr 'sprintf("**Observed prevalence of raised bog in the wet column of the climate grid, and where the future raised-bog cells land.** Training cells (current climate) with growing-season precipitation above 650 mm, by mean temperature of the warmest quarter (bio10). The last column is the share of the %s Norwegian raised-bog cells whose 2071--2100 climate falls in that cell of the grid; the other moisture columns together hold %s%% of them (full grid: S4 Table).", format(fr("raised bog", "Norway"), big.mark = ","), round(100 * (1 - fut_wet)))'
occ |>
  filter(gsp_band == WET) |>
  transmute(
    `bio10 (°C)` = sub("-", "–", bio10_band),
    Cells = n0(n_cells),
    `Raised bog` = n0(n_bog),
    Prevalence = sprintf("%.2f%%", prevalence_pct),
    `Share from Europe` = sprintf("%.0f%%", share_eu_pct),
    `Future raised-bog cells` = sprintf("%.1f%%", future_bog_pct)
  ) |>
  kable(align = "lrrrrr")
```

![**The climate the projection moves into, and what is observed there.** (A) Training
cells in the space of warmest-quarter temperature and growing-season precipitation
(15,000 cells drawn at random from each absence class; all `r n0(fr("raised bog"))` raised-bog cells), with the
`r n0(fr("raised bog", "Norway"))` Norwegian raised-bog cells at their 2071--2100 climate (red). Dotted lines are the
bands of panel B. (B) Observed prevalence of raised bog per band, with the number of raised-bog cells
over the number of training cells, and the share of the future raised-bog cells that land
in each band.](figures/fig_occupancy.png){#fig-occupancy}

## Cross-validated skill

Held-out skill depended on how much the training set resembled the production model
(@tbl-cv). Leave-one-partition-out folds, with training sets of the production model's size,
reached a macro AUC of `r f2(cvv("lopo", "all", "macro_auc"))` and recovered `r pc(cvv("lopo", "all", "recall_bog"))` of raised bogs at a false-positive rate of
`r f2(cvv("lopo", "all", "fpr_bog"))`; pairwise folds, trained on a median of `r n0(cvv("pairwise", "all", "median_n_train"))` cells (`r pc(cvv("pairwise", "all", "median_n_train") / n_all)` of the production set),
reached `r f2(cvv("pairwise", "all", "macro_auc"))` and recovered `r pc(cvv("pairwise", "all", "recall_bog"))`. The pairwise folds have both less training data and a
longer extrapolation distance, so we quote the leave-one-partition-out arm as the estimate of
production skill under climate transfer. Raised bogs were recovered on both data blocks
(`r pc(cvv("lopo", "NO", "recall_bog"))` in Norway, `r pc(cvv("lopo", "EU", "recall_bog"))` in Europe). Norwegian other peatland was recovered less well
(`r pc(cvv("lopo", "NO", "recall_otherpeat"))` against `r pc(cvv("lopo", "EU", "recall_otherpeat"))` in Europe; S3 Table), plausibly because mapped peatland is flatter in
Europe (`r pc(S("otherpeat_flat_share_EU"))` of European other-peatland cells have a slope below 1°, against `r pc(S("otherpeat_flat_share_NO"))` of
Norwegian ones) and slope is the most important predictor in the forest (S1 Table). This
weakness lies on the flat-versus-steep axis, not on the climatic axis the study turns on.

Beyond the temperature range of the training bogs the model was asymmetric. When the warmest
partition was held out, `r n0(cvv("lopo", "above", "n_test_rows"))` test cells were warmer than every training bog, none of them a
raised bog, and the model called `r pc(cvv("lopo", "above", "fpr_bog"))` of them raised bog (pairwise arm: `r pc(cvv("pairwise", "above", "fpr_bog"))`). The forest,
asked about a climate warmer than any raised bog it has seen, does not withdraw the raised-bog
label; it extends it. The projected loss reported below is therefore not an artefact of a model that
defaults to absence outside its training range.

```{r}
#| label: tbl-cv
#| tbl-cap: "**Cross-validated skill on partitions cut along bio10**, by arm (means over folds, all test cells). Recall is the share of raised-bog cells predicted raised bog; the false-positive rate is the share of non-bog cells predicted raised bog. Per-class recall, the G-mean and the split by data block and by position relative to the training bogs are in S3 Table."
cv |>
  filter(slice == "all") |>
  mutate(arm = factor(arm, c("lopo", "pairwise"))) |>
  arrange(arm) |>
  transmute(
    Arm = c(lopo = "Leave-one-partition-out", pairwise = "Pairwise")[as.character(arm)],
    Folds = n_folds,
    `Median training cells` = n0(median_n_train),
    `Macro AUC` = f2(macro_auc),
    `Recall raised bog` = f2(recall_bog),
    `FPR raised bog` = f2(fpr_bog)
  ) |>
  kable(align = "lrrrrr")
```

## Skill against warming beyond the training bogs

Along the signed temperature offset, the operating point moved more than discrimination did
(@fig-skill, S5 Table). Up to +`r f1(bins$axis_hi[5])` °C above the median training bog (bins 1--5) the pairwise models recovered
`r pc(min(ci$estimate[ci$metric == "recall_bog" & ci$bin <= 5]))`--`r pc(max(ci$estimate[ci$metric == "recall_bog" & ci$bin <= 5]))` of raised bogs at false-positive rates of `r f2(min(ci$estimate[ci$metric == "fpr_bog" & ci$bin <= 5]))`--`r f2(max(ci$estimate[ci$metric == "fpr_bog" & ci$bin <= 5]))`. In the
warm bin (`r sg(bins$axis_lo[WARM])` to `r sg(bins$axis_hi[WARM])` °C) recall rose to `r bci(WARM, "recall_bog")` and the false-positive rate to
`r bci(WARM, "fpr_bog")`, while the raised-bog AUC was at its lowest of the eight bins (`r bci(WARM, "auc_bog")`
against `r f2(min(ci$estimate[ci$metric == "auc_bog" & ci$bin < WARM]))`--`r f2(max(ci$estimate[ci$metric == "auc_bog" & ci$bin < WARM]))` elsewhere), though the intervals overlap. The warm bin
draws on only `r bins$n_folds[WARM]` folds and its intervals are wide. The forest asked about a climate
warmer than its training bogs thus extends the raised-bog label rather than withholding it,
consistent with the held-out warmest partition above, and ranks raised bogs somewhat less well.

The `r n0(S("n_lyngstad_polygons"))` mapped raised bogs sit today at a median offset of `r sg(S("lyngstad_offset_current_median"))` °C, inside bin
`r bin_of(S("lyngstad_offset_current_median"))`. Under the 2071--2100 climate they sit at a median offset of `r sg(S("lyngstad_offset_future_median"))` °C, inside the warm bin,
and `r pc(S("lyngstad_share_beyond_bog_edge"))` of the polygons lie beyond the warmest raised bog in the production training set
(`r sg(S("bog_edge_offset"))` °C). The skill the map can attach to them is the warm bin's.

![**Cross-validated skill as a function of warming beyond the training raised
bogs.** Held-out cells of the pairwise arm, binned on their warmest-quarter
temperature offset from the median of the raised bogs in that fold's training set (eight
bins with equal numbers of raised-bog cells; points at bin medians). (A) Share of true
raised-bog cells called raised bog (recall) and share of non-bog cells wrongly called raised
bog (false-positive rate), at the argmax operating point. (B) One-vs-rest AUC for raised bog
and macro AUC. Ribbons are 95% cluster-bootstrap intervals over the 20 folds. Thin dashed
lines: point estimates of the leave-one-partition-out arm, whose training sets are the
production model's size, for reference. Dashed vertical line: the warmest training bog. Grey
band: 10th--90th percentile of the 2071--2100 offset of the `r n0(S("n_lyngstad_polygons"))`
mapped raised bogs; their current offsets lie near zero.](figures/fig_skill_offset.png){#fig-skill}

## Projected change at the mapped raised bogs

Under current climate the model reproduces the mapped distribution: `r n0(tr_n(cur = "raised bog"))` of the `r n0(sum(trans$n_polygons))` polygons
(`r pc(tr_n(cur = "raised bog") / sum(trans$n_polygons))`) are predicted raised bog (@fig-projection (A, C)). Under the 2071--2100 climate the
raised-bog score falls at `r pc(S("lyngstad_share_declining"))` of the polygons (median change `r f2(S("lyngstad_pbog_change_median"))`), and `r n0(tr_n(cur = "raised bog", fut = "non-peat"))` of
the `r n0(tr_n(cur = "raised bog"))` predicted raised bogs are predicted non-peat and `r n0(tr_n(cur = "raised bog", fut = "other peatland"))` other peatland
(@tbl-transitions; @fig-projection (B, D); S1 Fig). The direction of the transition is the
point of the three-class response: the mapped raised bogs are not projected to become other
kinds of mire but to leave the peat-supporting climate altogether. Over mainland Norway as a
whole, the share of 250 m cells whose majority class is raised bog falls from `r pc(sh("current", "bog"), 1)` to
`r pc(sh("future", "bog"), 1)`, and other peatland from `r pc(sh("current", "otherpeat"))` to `r pc(sh("future", "otherpeat"))`.

<!-- TODO(JV): whether to report "newly suitable" area at all; notebook 2026-06-05 frames
the cold-flank transition (other peatland warming into the bog window) as a second
question. -->

```{r}
#| label: tbl-transitions
#| tbl-cap: !expr 'sprintf("**Predicted class of the %s mapped raised bogs under current and 2071--2100 climate** (argmax of the polygon-mean class probabilities).", format(sum(trans$n_polygons), big.mark = ","))'
trans |>
  transmute(
    `Predicted today` = cap1(class_current),
    `Predicted 2071–2100` = cap1(class_future),
    Polygons = n0(n_polygons),
    Share = sprintf("%.1f%%", share_pct)
  ) |>
  kable(align = "llrr")
```

![**Projected class probabilities over mainland Norway.** Raised-bog score under
(A) 1981--2010 and (B) 2071--2100 climate, and the majority class under (C) 1981--2010 and
(D) 2071--2100. Rasters aggregated to 2 km for display; outlines show the survey
footprint.](figures/fig_projection.png){#fig-projection}

## Reliability of the projection where it is read

By the multivariate dissimilarity index the future projection domain is not extreme: only
`r f1(rl("future", "pct_outside_aoa"), 1)`% of the future cells exceed the applicability threshold (`r f2(S("aoa_threshold"))`), mostly in the western
mountains, and `r f1(rL("future", "pct_outside_aoa"), 0)`% of the mapped-bog polygons do. That reading is
optimistic in a specific way: the index measures distance to any training cell, and the future
climate of the bogs is close to European non-peat and other-peatland cells, not to raised
bogs. On the signed temperature axis the same cells sit beyond the warmest training bog
(@fig-reliability (A)). The two layers say different things and both are part of the answer:
the projection is well supported by labelled data of the two other classes, and it is an
extrapolation with respect to raised bogs specifically.

The expected-skill layer (@fig-reliability (B)) carries the warm bin's recall over the
south-eastern and coastal lowlands and the cold bin's recall over the mountains and the north.
At the mapped raised bogs the expected recall is `r f2(rL("future", "exp_recall_bog"))` (`r f2(rL("future", "exp_recall_bog_lo"))`--`r f2(rL("future", "exp_recall_bog_hi"))`) under
the future climate against `r f2(rL("current", "exp_recall_bog"))` today, and the expected false-positive rate `r f2(rL("future", "exp_fpr_bog"))` against
`r f2(rL("current", "exp_fpr_bog"))`. The operating point the projection is read at favours calling raised bog,
which makes the non-peat verdict at these polygons stronger evidence than the same verdict
from a neutral operating point would be. The interval on the future recall is wide, however (`r f2(rL("future", "exp_recall_bog_lo"))`--`r f2(rL("future", "exp_recall_bog_hi"))`), so the verdict
is weaker than the point estimate suggests.

![**Reliability layers over the 2071--2100 projection domain.** (A) Warmest-quarter
temperature offset from the production model's training raised bogs. (B) Expected recall of
raised bog, read from the offset bins of @fig-skill; grey cells lie beyond the applicability
threshold of the dissimilarity index (`r f2(S("aoa_threshold"))`; `r f1(rl("future", "pct_outside_aoa"), 1)`% of cells) and carry no estimate.
Rasters aggregated to 2 km.](figures/fig_reliability.png){#fig-reliability}

# Discussion

TODO(JV): not drafted. Points the Results are written to support: (i) the model-free
occupancy table carries the conclusion and the model adds the map; (ii) the held-out warm
partition and the warm offset bin show the forest extends the raised-bog label into warmer
climates when the warm labels are withheld, which is the reply to "the model just defaults
to absence outside its range"; (iii) formation versus persistence: occurrence-based evidence
says which climates maintain raised bogs today, existing bogs are landforms with
self-maintaining hydrology, so the projection is a loss of climatic suitability as currently
expressed, not a disappearance date (notebook 2026-08-21; @joostenMiresPeatlandsEurope2017
p. 135); (iv) the warm evidence is European, southern and geographically clustered; (v) one
scenario and one climate model; (vi) the other-peatland weakness in Norway is a terrain-source
confound, not a climatic one; (vii) the absence-draw stratification on temperature shapes
observed prevalence across temperature bins (notebook 2026-10-05).

# Supporting information

::: {.supp}
## S1 Text. {#s1-text}

Data processing.

*European presences.* Of the good-quality raised-bog sites in region IV, `r n0(aud_col("large, discarded", "polygons"))` large sites without
a plot record (`r n0(aud_col("large, discarded", "area_km2"))` km^2^) were not used as presences. The loss costs power but not climatic
coverage: their median warmest-quarter temperature lies between that of the retained small
sites and that of the retained large sites with plots, and the small sites span the widest
temperature range of the three sets (polygon values on the 5 km working grid, which is used
only to define the domain and the temperature strata, never as a predictor surface).

*Label counts.* Norway: of the absence candidates, 1.32 million were non-peat, 61,850 other
peatland, 177,860 mixed and 197,190 without an AR50 value (before the land-use screen).

*Predictor extraction.* Climate and palaeoclimate layers were projected from their native 30
arc-second grid to a 250 m grid for Norway, which is both the training surface for the Norwegian
block and the projection domain. No European surface was built: predictor values for the
European cells were extracted at 250 m cell centres by bilinear interpolation from the same
native grids (terrain from the 250 m European terrain raster), so that both blocks are bilinear
samples of the same native field. Growing-degree, growing-season-temperature and snow variables
that CHELSA leaves undefined where the threshold is never crossed were set to zero where all
other layers had data. Growing-season precipitation is stored without a no-data tag, so its
no-data code was masked explicitly before resampling; the pipeline stops if any value implies
a growing-season precipitation above 100,000 mm, and no training or projection cell carries
the code in the results reported here. Elevation outside Norway was fetched at about 90 m with
the elevatr package and aggregated to 250 m before slope was computed.
:::

::: {.supp}
## S1 Table. {#s1-table}

Predictors.

The 43 predictors with their source, whether they change under the scenario (and so enter
the dissimilarity metric), the mean projected shift in standard deviations, and permutation
importance in the production forest (all classes, and for the raised-bog class). Built by
ms/make_figures.R as ms/tables/tblS_predictors.csv.
:::

::: {.supp}
## S2 Table. {#s2-table}

Partitions.

Range on bio10 and class composition of the five cross-validation partitions
(ms/tables/tbl_partitions.csv).
:::

::: {.supp}
## S3 Table. {#s3-table}

Cross-validated skill in full.

```{r}
#| label: tbl-cv-full
#| echo: false
lab_slice <- c(
  all = "all",
  NO = "Norway",
  EU = "Europe",
  above = "warmer than all training bogs",
  below = "colder than all training bogs"
)
f2na <- function(x) ifelse(is.na(x), "--", f2(x))
cv |>
  filter(slice != "inside") |>
  mutate(
    arm = factor(arm, c("lopo", "pairwise")),
    slice = factor(slice, names(lab_slice))
  ) |>
  arrange(arm, slice) |>
  mutate(
    env = slice %in% c("above", "below"),
    Arm = ifelse(
      duplicated(arm),
      "",
      c(lopo = "Leave-one-out", pairwise = "Pairwise")[as.character(arm)]
    ),
    `Test cells` = unname(lab_slice[as.character(slice)]),
    Folds = n_folds,
    `Median training cells` = n0(median_n_train),
    `Macro AUC` = f2na(macro_auc),
    `Macro G-mean` = ifelse(env, "--", f2na(macro_gmean)),
    `Recall raised bog` = ifelse(env, "--", f2na(recall_bog)),
    `Recall other peatland` = f2na(recall_otherpeat),
    `Recall non-peat` = f2na(recall_nonpeat),
    `FPR raised bog` = f2na(fpr_bog)
  ) |>
  select(
    Arm,
    `Test cells`,
    Folds,
    `Median training cells`,
    `Macro AUC`,
    `Macro G-mean`,
    `Recall raised bog`,
    `Recall other peatland`,
    `Recall non-peat`,
    `FPR raised bog`
  ) |>
  kable(align = "llrrrrrrrr")
```

Means over folds. The two envelope rows (test cells warmer or colder than every training
bog on bio10) contain no raised bogs, so raised-bog recall and the G-mean are undefined there
and the false-positive rate is the informative column.
:::

::: {.supp}
## S4 Table. {#s4-table}

Observed prevalence of raised bog across the full climate grid.

```{r}
#| label: tbl-occupancy-full
#| echo: false
occ |>
  transmute(
    `bio10 (°C)` = sub("-", "–", bio10_band),
    `gsp (mm)` = sub("-", "–", gsp_band),
    Cells = n0(n_cells),
    `Raised bog` = n0(n_bog),
    Prevalence = sprintf("%.2f%%", prevalence_pct),
    `Share from Europe` = sprintf("%.0f%%", share_eu_pct),
    `Future raised-bog cells` = sprintf("%.1f%%", future_bog_pct)
  ) |>
  kable(align = "llrrrrr")
```
:::

::: {.supp}
## S5 Table. {#s5-table}

Held-out skill by bin of warmest-quarter temperature offset from the training raised bogs
(pairwise arm). Intervals are 95% cluster-bootstrap intervals over the 20 folds.

```{r}
#| label: tbl-bins
#| echo: false
ci |>
  filter(metric %in% c("recall_bog", "fpr_bog", "auc_bog", "macro_auc")) |>
  mutate(
    cell = ifelse(
      is.na(lower),
      f2(estimate),
      sprintf("%s (%s–%s)", f2(estimate), f2(lower), f2(upper))
    )
  ) |>
  select(bin, n, n_bog, n_folds, axis_lo, axis_hi, metric, cell) |>
  pivot_wider(names_from = metric, values_from = cell) |>
  transmute(
    Bin = bin,
    `Offset range (°C)` = sprintf("%+.1f to %+.1f", axis_lo, axis_hi),
    Cells = n0(n),
    `Raised bogs` = n0(n_bog),
    Folds = n_folds,
    `Recall raised bog` = recall_bog,
    `FPR raised bog` = fpr_bog,
    `AUC raised bog` = auc_bog,
    `Macro AUC` = macro_auc
  ) |>
  kable(align = "rlrrrllll")
```
:::

::: {.supp}
## S1 Fig. {#s1-fig}

Change at the `r n0(S("n_lyngstad_polygons"))` mapped raised bogs.

![(A) Polygon-mean raised-bog score under current against future climate (dotted line: no
change). (B) The polygons coloured by their 2071--2100 raised-bog
score.](figures/fig_lyngstad_change.png)
:::

# Acknowledgments

TODO(JV).
